### 2023-01-16 Paper/MSEPM-Manuscript.tex
First Draft

\documentclass{article}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{caption}
% \usepackage{psfig}
%\usepackage{epic}
%\usepackage{eepic}
\usepackage{colordvi}
%\usepackage{wrapfig}  
\usepackage{afterpage}
%\usepackage{placeins}
\usepackage{graphics,color}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{placeins}
\usepackage{longtable}
\usepackage{color}      % Need the color package
\usepackage{epsfig}
\usepackage{subfiles}
\usepackage[style=nature]{biblatex}
\usepackage{float}
\usepackage{lineno}

\newcommand{\E}{\mathrm{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\cherry}[1]{\ensuremath{\langle #1 \rangle}}
\newcommand{\eopf}{\framebox[6.5pt]{} \vspace{0.2in}}

%\def\l{{\lambda }}
\def\e{{\varepsilon}}
\def\a{{\alpha }}
\def\S{{\Sigma }}
\def\s{{\sigma }}
\def\M{{\mathcal M}}
\def\X{{\mathcal X }}

\bibliography{MSEPM-Manuscript}

\begin{document}

\large{\bf{The Multi-State Epigenetic Pacemaker enables the 
identification of combinations of factors that influence DNA methylation}}\\ \\
\noindent{\small Colin Farrell$^{1,3}$, Kalsuda Lapborisuth$^{1}$, 
                 Sagi Snir$^{2}$,\\ and Matteo Pellegrini$^{1,3}$\\ \\
\noindent{\footnotesize 
$^1$Dept. of Molecular, Cell and Developmental Biology; \\ University of
California, Los Angeles, CA 90095, USA;;\\
\noindent
$^2$Dept. of Evolutionary Biology, University of  Haifa, Israel;\\
$^3$Corresponding Authors; colinpatfarrell@g.ucla.edu, matteop@mcdb.ucla.edu 
}

\begin{linenumbers}
\begin{center}\rule{0.9\linewidth}{0.5pt}\end{center}

Epigenetic clocks, DNA methylation based predictive models of chronological age, are often utilized to study aging 
associated biology. Despite their widespread use, these methods do not account for other factors that also 
contribute to the variability of DNA methylation data. For example, many CpG sites show strong sex-specific 
or cell type specific patterns that likely impact the predictions of epigenetic age.  To overcome these 
limitations, we developed a multidimensional extension of the Epigenetic Pacemaker, the Multi-State Epigenetic 
Pacemaker (MSEPM). We show that the MSEPM is capable of accurately modeling multiple methylation associated factors 
simultaneously, while also providing site specific models that describe the per site relationship between methylation 
and these factors. We utilized the MSEPM with a large aggregate cohort of blood methylation data to construct models 
of the effects of age, sex and cell type heterogeneity on DNA methylation.  We found that these models capture a 
large faction of the variability at thousands of DNA methylation sites.  Moreover, we found modeled sites that are 
primarily affected by aging and no other factors.  Among these, those that lose methylation over time are enriched 
for CTCF transcription factor chip peaks, while those that gain methylation over time are enriched for REST 
transcription factor chip peaks. Both transcription factors are associated with transcriptional maintenance and 
suggest a general dysregulation of transcription with age that is not impacted by sex or cell type heterogeneity. 
In conclusion, the MSEPM is capable of accurately modeling multiple methylation associated factors and the models 
produced can illuminate site specific combinations of factors that affect methylation dynamics.

\begin{center}\rule{0.9\linewidth}{0.5pt}\end{center}
\section{Introduction}

DNA methylation, the addition of a methyl group to the fifth carbon of the cytosine pyrimidine ring,  is 
associated with the topological organization of the cellular genome, gene expression and the state of 
a cell. Within a population of cells the methylation pattern at certain sites can change predictably 
with the age of the individual from which the cells are drawn. This predictable nature of DNA 
methylation has led to the development of accurate DNA methylation based predictive models for 
age and health, termed epigenetic clocks. The difference between the predicted and the expected 
epigenetic age given an individual's chronological age has been interpreted as a measure of age 
acceleration\cite{Horvath2018-ia}, and has been associated with mortality\cite{Perna2016-pi,Marioni2015-sn} 
and other adverse health outcomes\cite{Dugue2018-ad,Huang2019-hf,Armstrong2017-vg,Chuang2017-nk,Horvath2015-af}.

However, epigenetic clocks suffer from several limitations that limit the interpretability of their 
predictions and the underlying mechanisms. Epigenetic clocks are generally trained by using penalized regression 
based methods that attempt to minimize the difference between the predicted and observed value of  age. As a result, 
as the error between predicted and observed age is decreased, the associations between age acceleration and mortality 
disappears\cite{Zhang2019-br}. Second generation epigenetic clocks attempt to resolve this issue by fitting a measure 
of human health, rather than age, and as a result these clocks are generally more sensitive to individual health 
status \cite{Lu2019-lg,Levine2018-en,Belsky2020-ha}. However, while the response variable is modified in these 
clocks the method used to fit the clock is largely the same. Epigenetic clocks are generally trained using 
regularized regression models, where the likelihood is maximized by minimizing the difference between the 
observed and predicted response variable subject to the elastic net penalty,$\lambda_1$ and $\lambda_2$. 
Methylation sites that increase model error and are influenced by other relevant factors such as smoking or obesity, 
may be discarded during model fitting, thus limiting the ability of this approach to account for the effects of these 
extraneous factors on epigenetic aging.

As an alternative to penalized regression based methods we previously developed an evolutionary based model for 
epigenetic dynamics, the Epigenetic Pacemaker (EPM)\cite{Farrell2020-bn,Snir2016-dv}. The EPM attempts to 
minimize the difference between observed and predicted methylation values amongst a collection of sites through the 
implementation of a conditional expectation maximization algorithm\cite{Snir2020-tc}. Under the EPM the observed 
methylation status of a collection of sites is modeled linearly with respect to an input factor of interest, such 
as age. A hidden epigenetic state, that is related to the initial factor, but not necessarily linearly, is learned 
through the course of model fitting. The EPM can capture the non-linear relationship between methylation and 
age\cite{Snir2019-ii} and outputs an interpretable model for each site. However, both the EPM and regression 
based methods suffer from the same limitation, which is that they are limited to a single trait predicted by, 
or used to model, observed methylation patterns. In reality, the observed methylation landscape is likely impacted 
by a variety of factors that act simultaneously to produce the observed methylome of an individual.

To overcome this limitation, we have developed a multidimensional extension of the EPM, the 
Multi-State Epigenetic Pacemaker (MSEPM).  We show that the (MSEPM) can accurately model  site specific 
methylation variation driven by several factors, and given a trained model, accurately predict the values of the 
factors associated with an individual's observed methylation profile in both simulated methylation datasets and a 
large aggregate blood tissue methylation dataset. Importantly, as factors that explain the observed methylation 
profile of an individual are added to the model the ability to model the factors and methylation values improves. 
Additionally, we show that sites with similar associations to modeled factors cluster together and are enriched 
for specific transcription factors.

\section{Results}

\subsection{Multi-State Epigenetic Pacemaker Model}

The MSEPM attempts to describe the observed methylation status at any single methylation site as a linear combination 
of methylation associated factors specific to an individual, termed epigenetic factors. Under this model epigenetic 
factors are related to observable individual factors, such as chronological age, sex and cell types, but may be 
transformed relative to observable factors. The epigenetic age factor, for example, often has a non-linear 
relationship with the observed age \cite{Snir2019-ii}. The MSEPM learns the appropriate transformation during model 
fitting to describe the observed methylation status linearly in terms of the epigenetic age factor 
(and not linearly with age). Given a site $i$ and individual $j$ the observed methylation status can be 
modeled as $\hat{m_{i,j}} = r^0_i + \sum^n_{k=1} f_{j,k} r_{i,k}$, where $k$ epigenetic factors, $f_{j,k}$, 
are weighted by $k$ site specific parameters, $r_{i,k}$, and offset by a sites specific intercept term, $r^0_i$. 
Site parameters, $r_{i,k}$ and $r^0_i$, are characteristic of the site and shared amongst all individuals while 
epigenetic factors, $f_{j,k}$, are characteristic of an individual and are the same across all sites for that 
individual. As a consequence, observed methylation differences between any two individuals are dependent on 
individual epigenetic factors $f_{j,k}$. In practice, the observed methylation value is also dependent on a 
normally distributed error term 
$\epsilon_{i,j}$, $\hat{m_{i,j}} = r^0_i + \sum^n_{k=1} s_{j,n} r_{i,n} + \epsilon_{i,j}$.

Given an input matrix $\hat{M} = [m_{i,j}]$ the objective of the MSEPM is to find the optimal values of 
$r_{i,k}$ and $f_{k,j}$ that minimize the  residual sum of square (RSS) error, 
$\epsilon_{i,j}^2 = (\hat{m_{i,j} – r^0_{i} - \sum^n_{n=1} f_{j,k} r_{i,k} })^2$. This is 
accomplished through the implementation of a conditional expectation maximization algorithm. The 
maximum likelihood (ML) values of $r_{i,k}$ and $r^0_i$ for each $i$ methylation site can be solved 
using ordinary least squares (OLS) regression. Given the ML $r_{i,k}$ site coefficients each $k$ epigenetic 
factor is then updated by fixing the site coefficients and updating $s_{k,j}$ by minimizing the RSS across 
all $i$ sites using gradient descent, $f_{k,j}^{n+1} = f_{k,j} - \lambda \nabla F(f_{k,j})$, where $\lambda$ 
is a specified learning rate. The optimization is accomplished by alternating between optimizing $r_{k,i}$ and 
$f_{k,j}$ until the reduction in sum of the site RSS is below a specified threshold between iterations or a set 
number of interactions is reached. Importantly, while the ML values of $f_{j,k}$ are by definition linear with 
the methylation status at any site, the original input factors for $f_{j,k}$ may not be.

Provided a trained MSEPM model and an unobserved methylation matrix, epigenetic factors are estimated by 
calculating each independent OLS for solution all $i$ sites given the $r_{i,k}$ coefficients set for the 
respective input factor.  These epigenetic factors can then be used to find the expected methylation value using the 
trained individual site models where $E[m_{i,j}] = r_{i,0} + F_{j} \dot R_{i}$.

\subsection{Simulated Methylation Associated Factors}

We implemented a simulation framework that extends the MSEPM model where the methylation status at site $i$ for an 
individual $j$ is described by a weighted sum of epigenetic factors, 
$\hat{m_{i,j}} = r^0_i + \sum^n_{l=1} f_{k,j} r_{i,k} + \epsilon_{i,j}$. As shown in our 
previous work \cite{Snir2019-ii} the association between a methylation associated input factor and methylation 
status is not necessarily linear. To account for non-linear associations, $f_{k,j}$, have the 
form $f_{j,k} =  q_{j,k}^{\gamma_{k}}$, where $\gamma_{k}$ is the transformation between a factor of 
magnitude $q_{n,j}$ and the epigenetic factor. In practice the value of the $f_{k,j}$ is often unknown and the 
association between methylation status and $f_{k,j}$ is inferred through $q_{n,j}$. We simulated individuals 
whose methylation is determined by four factors and their associated epigenetic factors: a uniform factor 
approximating age with a non-linear association with methylation status 
($q \sim\mathcal{U}(0,100), s_{Age}=q^{0.5}$, Figure 1A-B), a binary trait resembling sex, 
linearly associated with methylation status ($q \sim B(1,.5), s_{Sex}=q$, Figure 1C-D), a continuous normal 
(CN) phenotype resembling a cell type with a linear association with methylation 
status $(q \sim\mathcal{N}(1, 0.1), s_{CN}=q$, Figure 1E-F), and a continuous exponentially (CE) distributed trait 
resembling obesity with a linear association with methylation status 
$(q \sim \frac{1}{20}e^{-x/20}, s_{CE}=q, Figure G-H)$. Methylation sites were simulated by first randomly 
setting the dynamic range of the methylation site, ($ -1< \delta < 1$), a site intercept, $r^0_i$, and the 
site error, $(\sigma_i \sim \mathcal{U}(0.025, 0.05)$). Simulated methylation sites were associated with a 
combination of zero, one, or multiple epigenetic factors. Rates for sites associated with multiple factors were 
set by sampling from a uniform distribution. Site rates were normalized to describe the dynamic range of the 
simulated methylation site unless the rates for all phenotypes were zero.

We simulated 90 methylation sites (Figure 1I). We then evaluated the MSEPM model as follows. 
We simulated 1000 samples with the four epigenetic factors described above. We then simulated methylation values
 using the simulated site rates. Simulated samples were then split for training ($n=500$) and testing ($n=500$).
  MSEPM models were then fitted using the values of the input factors, $q_{k,j}$, as the initial guess for
   $f_{k,j}$. We generated 1000 simulated datasets and fit MSEPM models using four combinations of input 
   factors (Age, Age-Sex, Age-Sex-CN, Age-Sex-CN-CE). Within each simulation, epigenetic state predictions and 
   methylation site predictions were made for all testing samples. All models captured the nonlinear association 
   between simulated age and methylation (S. Figure 1). As the number of factors in the model is increased the mean 
   absolute error (MAE) between the predicted epigenetic states and the simulated epigenetic factors decreases 
   (Figure 2A). Importantly, to accurately assess simulated age it is necessary to account for the influence of 
   the other simulated factors (Sex, CN, CE). The MAE between the predicted and simulated methylation values decreases 
   as simulated factors are added to the model, and accurately assessing the methylation status of a simulated site 
   requires that the factor associated with the methylation status at the site is included in the model (Figure 2A).

The MSEPM model generated using all four simulated factors can capture the relative magnitude of the simulated 
site-specific rates (Figure 2C-F).  However, the model has difficulty capturing the exact relationship between the 
simulated factors (age, CN and CE) and the inferred factors (Figure 2C, E-F). This is likely due to limitations of 
the model at capturing nonlinear methylation association and a limited training range for normally and exponentially 
distributed traits. Regardless, the four-factor model can accurately predict the simulated methylation value 
(Figure 2 D) and site intercept (Supp. Figure 1G). We also assessed the model robustness to variation in the 
number of samples and sites used for model training by randomly selecting a reduced subset of samples or sites for 
model training. MSEPM models trained with age, sex, CN, and CE can accurately assess all simulated phenotypes with 
few samples and sites (Supp Figure2 B-E).

\subsection{Blood MSEPM Model}

We utilized a large aggregated dataset composed of Illumina 450k array data from 17 publicly available 
datasets \cite{Ventham2016-qj,Demetriou2013-wb,Polidoro2013-xv,Johansson2013-of,Arloth2020-lo,Liu2013-dg,Soriano-Tarraga2016-uq,Chuang2017-nk,Zannas2019-me,Kurushima2019-pe,Voisin2015-lh,Tan2014-sg,Tserel2015-ro,Butcher2017-oz,Dabin2020-iw,Marabita2013-cj,ValleG2019-xi} 
deposited in the Gene Expression Omnibus \cite{Barrett2012-gu} (GEO) generated from blood derived samples 
(whole blood, peripheral blood lymphocytes, and peripheral blood mononuclear cells). All datasets were processed 
through a unified pipeline from raw array intensity data (IDAT) files using minfi (Aryee et al., 2014). Sex and 
blood cell type abundance predictions were made for each as previously described\cite{Houseman2012-rr,Aryee2014-ky}. 
The aggregate data spanned a wide age range (0.0  - 99.0 years, Figure 3A), contained more predicted females 
($n=3392$) than males ($n=2295$, Figure 3B) and produced reasonable cell type abundance estimates (Figure 3C).

We trained MSEPM models using data assembled from four GEO 
series\cite{Johansson2013-of,Liu2013-dg,Butcher2017-oz,Damaso2020-gd}  ($n=1605$) with samples spanning a wide age 
range (0.01 - 94.0 years). The samples were randomly split into training ($n=1203$) and validation ($n=402$) sets 
stratified by age. Training set blood cell type abundance estimates were used to train a principal component analysis 
(PCA) model which was then used to calculate cell type PCA estimates for the validation and testing sets. The first 
cell type principal component (CT-PC1) is largely driven by the relative abundance of granulocytes (Figure 3D), while 
the second PC (CT-PC2) captures relative differences in the abundance of differentiated lymphocytes (Figure 3D).  
Methylation values for all samples were quantile normalized by probe type \cite{Horvath2013-sk} using the median 
site methylation values across all training samples for each methylation site.

Methylation sites were selected for modeling with MSEPM if the sites were highly correlated with 
age ($n=276$) , sex ($n=49$), CT-PC1 ($n=120$), CT-PC2 ($n=116$) or a combination of factors ($n=238$), 
a total of 778 unique sites. Four MSEPM models were fit using the selected sites with four combinations of input 
factors (Age, Age Sex, Age Sex CT-PC1, and Age Sex CT-PC1 CT-PC2). Training sample factors were min-max scaled between 
0 and 1 before model training. The association between the fit epigenetic factor predictions against the input modeled 
factors was assessed by fitting a trendline between epigenetic state predictions and scaled continuous input factors 
using the state prediction made for the MSEPM model trained with all four input factors. Performance of the MSEPM 
model was then evaluated using the testing samples ($n=4,082$). The performance of the MSEPM largely closely 
resembles the simulation results. All four MSEPM models capture the nonlinear relationship between age and 
methylation status (Supp. Figure 6). The epigenetic state prediction associated with age improves as the underlying 
methylation data are more fully explained through the addition of epigenetic factors (Supp. Figure 6). The MSEPM model 
fit with Age, Sex, CT-PC1 and CT-PC2 can accurately model the associated epigenetic state for each factor (Figure 4 A-D) 
and accurately predicts the methylation levels at individual sites ($R^2=0.935$, $MAE=0.035$, Figure 4 E). 
The trained MSEPM produces a collection of methylation site models that can help explain the association between 
modeled factors and methylation status.

\subsection{Analysis of chromatin regulators of site clusters}

We evaluated the relationship between sites that are influenced by age, sex, CT-PC1 or CT-PC2  and potential 
regulatory factors by performing overlap enrichment analysis of these sites with transcription factor chromatin 
immunoprecipitation peaks present in the ENCODE V4 \cite{ENCODE_Project_Consortium2012-oe,Davis2018-ha} release. We 
first identified sites with similar coefficients of epigenetic factors through hierarchical clustering. Briefly, 
sites were clustered using Ward's method by the euclidean distance between site regression coefficients that were 
normalized by the standard deviation of the methylation values amongst the training samples. The resulting tree was 
cut at a height of 18 to produce 10 distinct clusters with clear associations to the modeled factors (Figure 5A).

The site clusters largely conform to underlying biological expectations. Cluster one contains sites that are wholly 
associated with sex status and localized to the X chromosome (Supp. Table 1) and is enriched for peaks of 
transcription factors associated with sex specific regulation such as MAZ \cite{Lopes-Ramos2020-ex}. Clusters nine and 
ten contain sites whose methylation status is largely driven by CT-PC1, and are enriched for transcription factors 
associated with granulocyte development (CEBPB, CEBPA, EP300, ETV6)\cite{Theilgaard-Monch2022-zw} \cite{Guerzoni2006-ii}.  
Similarly, clusters two, five and eight are associated with CT-PC2 and are enriched for transcription factor peaks 
associated with immune development (ZBED1, ETV6, FOSL2, FOS, TBX21). Clusters four and six are associated with loss of 
methylation with age. Cluster six is highly enriched for CTCF binding sites; CTCF is known to increase at sites where 
methylation is lost during aging \cite{Tharakan2020-pj}. Cluster four is enriched for STAT3 whose activation during 
exercise is age dependent \cite{Trenerry2008-kj,Mohamed2020-he}.  Cluster seven is associated with the accumulation 
of methylation with age and is enriched for immunoprecipitation peaks for aging associated transcription factors 
SMAD4 and RE1-Silencing Transcription Factor (REST). SMAD4 encodes a protein involved in the transforming growth 
factor beta (TGF-$\beta$) signaling pathway. Age related dysregulation of TGF-$\beta$ has been linked to reduced skeletal muscle 
regeneration\cite{Paris2016-fo,Carlson2009-uz} and SMAD4 polymorphisms are associated 
with longevity \cite{Carrieri2004-by}. REST is a transcriptional repressor of neuron specific genes in non-neuronal 
cells \cite{Chong1995-dj,Coulson2005-pb}. REST expression is upregulated in aged prefrontal cortex tissue and the 
absence of REST expression is associated with cognitive impairment \cite{Lu2014-dz} and 
cellular senescence in neurons \cite{Rocchi2021-od}.

\section{Discussion}

Epigenetic clocks are widely used tools to study human aging and health. Despite their widespread use, the 
biological interpretation of model parameters is limited. A methylome is influenced by many different biological 
processes occurring simultaneously over time that may differ among individuals. Epigenetic clocks, while producing 
accurate predictions of age, attempt to capture this complexity through a single dependent variable. Additionally, 
the penalized regression based methods used to fit most epigenetic clocks select sites that minimize, or regress out,
the influence of other factors and omit groups of sites that are correlated. To overcome these limitations, here we 
propose a multidimensional extension of the EPM model, the MSEPM.

In contrast to previous methods, the MSEPM aims to simultaneously model the effect of multiple factors on the 
methylome.The simulation and blood MSEPM models show that concurrently modeling age, cell type composition and sex 
can minimize  model residuals when compared with the MSEPM model fit with age only. The residual of the age only 
model is often interpreted as a measure of age acceleration. When multiple methylome associated traits are modeled 
simultaneously this residual can be explained directly by other factors and the association between the methylome 
and a trait of interest can be inferred.

Additionally, the individual methylation site linear models fit as part of the MSEPM optimization can provide 
information about the relationship between modeled factors and site specific biology.  To this end, we find that 
the blood MSEPM model conforms to expected biology. Sites with a strong sex association localize to the X chromosome 
and sites associated with cell types are enriched for transcription factors associated with the development of 
immune cells. Sites that are primarily affected only by age in the blood MSEPM model are of particular interest. 
As others have previously described, sites that progressively lose methylation over time are strongly enriched for 
CTCF \cite{Han2020-zj,De_Lima_Camillo2022-lu}, while in the blood MSEPM model sites that gain methylation over time 
are enriched for REST. The association between age and REST is well established, but the association between gain of 
methylation and age has previously not been reported. Both transcription factors are associated with the maintenance 
of transcriptional profiles and suggest that there is a general dysregulation of maintenance of cell specific 
chromosomal organization over time.

In conclusion, we introduced a multi-dimensional extension of the Epigenetic Pacemaker, the MSEPM. The MSEPM is 
capable of accurately modeling multiple methylation associated factors simultaneously. This paradigm can elucidate 
the site specific regulation underpinning methylome dynamics. The MSEPM is available under the MIT license at
 https://github.com/NuttyLogic/MultistateEpigeneticPacemaker.

\section{Methods}

\subsection{MSEPM}

MSEPM model training begins with an  input matrix of $i$ methylation sites for $j$ individuals 
and $k$ input traits for $j$ individuals. The goal of the MSEPM is to minimize the difference 
between the observed and predicted methylation values at any of the $i$ input methylation sites as 
explained by the $k$ input traits. The MSEPM model is fit as follows:
\begin{itemize}

    \item{Fit $i$ ordinary least squares regression models with the 
    form $\hat{m_{i,j}} = r^0_i + \sum^n_{k=1} s_{j,k} r_{i,k} + \epsilon_{i,j}$ to estimate site 
    specific parameters $r_i$ shared amongst all individuals. }
    \begin{itemize}
        \item{The $k$ untransformed input traits are used as the initial guess of $S_{j,k}$ 
        for the first model iteration. }
    \end{itemize}
    \item{Update $S_{j,k}$ to minimize the RSS cost 
    function, $J(k) = \frac{1}{n}\sum_{i=1}^n(\hat{m_{ij}} - (r^0_i + R_i S_j))^2)$, using 
    gradient descent with fixed $r_{i}$ site parameters.}
        \begin{itemize}
            \item{$\nabla J(k)$}
            \begin{itemize}
                \item{ $\frac{d}{dk} = \frac{2}{n} \sum_{i=1}^n (\hat{m_{ij}} - (r^0_i + R_i S_j)) r_{i,k}$}
                \item{$s^1_{j,k} = s^0_{j,k} - \eta * \frac{d}{dk}$, $\eta=$ learning rate}
            \end{itemize}
        \end{itemize}
    \item{Repeat steps 1-2 until the reduction in sum of site RSS between iterations is $\leq$ 
    specified threshold or a set number of iterations has been reached. }
\end{itemize}

\subsection{Simulation Framework}

Under the MSEPM formulation a single site can be described linearly where the observed methylation value is 
dependent on a weighted sum of $n$ epigenetic states, 
$\hat{m_{i,j}} = r^0_i + \sum^n_{n=1} p_{j,n} r_{i,n} + \epsilon_{i,j}$, where $p_{j,n}$ is an individual 
trait associated with the methylation status at site $i$. Using this formulation we implemented a simulation 
framework in python with only numpy \cite{Harris2020-yb} as a dependency.  The details of the approach are as follows:

\begin{itemize}
    \item{Individuals are simulated by setting individual traits with the following form 
    $p_{n,j} = Age_j^{\gamma_{n}} q_{n,j}$ Where the simulated individual trait $p_{n,j}$ is 
    dependent time and exposure, $q_{n,j}$, over time. Under this formulation traits may have a 
    nonlinear association with the simulated methylation status as mediated by time. The form of  $Q[q]$ varies 
    between simulated traits as shown above. Simulated MSEPM models are fit using $q_{n,j}$ as input for and the 
    association between $p_{n,j}$ is learned through model fitting.$p_{n,j} = Age_j^{\gamma_{n}} q_{n,j}$ Where 
    the simulated individual trait $p_{n,j}$ is dependent time and exposure, $q_{n,j}$, over time. Under this 
    formulation traits may have a nonlinear association with the simulated methylation status as mediated by time. 
    The form of  $Q[q]$ varies between simulated traits as shown above. Simulated MSEPM models are fit
    using $q_{n,j}$ as input for and the association between $p_{n,j}$ is learned through model fitting.}
    \item{Methylation site profiles are then simulated by setting the dynamic range of possible methylation values 
    and the error profile of the methylation site. The simulated methylation values are bounded by 0 and 1, and 
    represent the proportion of methylated bases to the total number of observed cytosines at any site. The 
    dynamic range is set by first initial methylation value $m_0 ~ \mathcal{beta}(.2,.2)$ and then setting a 
    target methylation value $m_t$ where the dynamic range is $\delta_i = m_t - m_o$. The value of $m_t$ is 
    set conditionally to ensure dynamic range is always larger than some specified threshold ,$\theta$, where 
    $ \theta \leq|\delta| \geq$. $m_0 ~ \mathcal{beta}(.2,.2)$ and then setting a target methylation value 
    $m_t$ where the dynamic range is $\delta_i = m_t - m_o$. The value of $m_t$ is set conditionally to 
    ensure dynamic range is always larger than some specified threshold ,
    $\theta$, where $ \theta \leq|\delta| \geq$. }
    \item{Methylation sites profiles are then randomly selected and associated with a weighted set of simulated 
    traits. The weighted traits are then normalized so the input combination of traits describes the dynamic 
    range of the simulation site, $\delta$. If no weights are provided the individual methylation sites are 
    simulated by returning $m_0 + N(\mu, \sigma)$. $\delta$. If no weights are provided the individual 
    methylation sites are simulated by returning $m_0 + N(\mu, \sigma)$. }
\end{itemize}

\subsection{Blood MSEPM Model Training}

Methylation sites with an absolute pearson correlation coefficient greater than 0.7, 0.995, 0.92 and 0.64 between 
age $(n=276)$, sex $(n=49)$, cell type PC1 $(n=120)$ and cell type PC2 $(n=238)$ and site methylation values 
respectively among training samples samples($n=1203$) were included in the model. Additionally, sites with a 
sum of all absolute pearson coefficients greater than 1.8 were also included $(n=238)$ for a total of 778 
methylation sites. Min-max, (0-1), scalers were fit using the training input features. Validation and testing 
sample features were transformed with the trained scalers, note age was min-max scaled on a range from 0-100 
years. MSEPM models were trained with a learning rate of 0.01 with an iteration limit of 200.

\subsection{Blood MSEPM Model Cluster Transcription Factor Overlap Analysis}

A custom transcription factor reference set was built from all ENCODE V4 transcription factor chromatin 
immunoprecipitation \cite{ENCODE_Project_Consortium2012-oe,Davis2018-ha} (release 1.4.0 - 2.1.2)  
irreproducible discovery rate narrow bed peaks, which contains peaks with high rank consistency between 
replicates, that were not audited for non-compliance or errors. GRCh38 region coordinates were 
lifted to GRCh37 coordinates using liftOver \cite{Hinrichs2006-oq}. The overlap reference contains 714 
transcription factor targets 1621 accession IDs (table …).

Blood MSEPM model site hierarchical clustering was performed as follows. Individual methylation site coefficients 
were first normalized by the standard deviation of methylation values of the site among the training samples, 
$r_{i,n} / \sigma_{i}$. A distance matrix was then created by taking the Euclidean distance between the 
normalized site model coefficients. Sites were then clustered using Ward's method which seeks to minimize 
within cluster variance by minimizing the increase in the error sum of squares (ESS) through successive 
cluster fusions.  Clusters label by tree cutting at a height of 18. All clustering analysis was carried out 
using SciPy v1.6.3 \cite{Virtanen2020-mm}.  $r_{i,n} / \sigma_{i}$. A distance matrix was then created by 
taking the Euclidean distance between the normalized site model coefficients. Sites were then clustered using 
Ward’s method which seeks to minimize within cluster variance by minimizing the increase in the error sum of 
squares (ESS) through successive cluster fusions.  Clusters label by tree cutting at a height of 18. All 
clustering analysis was carried out using SciPy v1.6.3 (Virtanen et al. 2020).

Transcription factor enrichment analysis was performed with  LOLA \cite{Sheffield2016-wg} which assesses the genomic 
region set overlap between a set of query regions and a set of reference regions, within a specified shared 
background set, using Fisher's exact test. Overlap analysis was performed for sites within a cluster against 
the ENCODE V4 reference region (1BP minimum overlap) using all sites assayed with 
Infinium HumanMethylation450K BeadChip as background.

\subsection{Analysis Environment}

Analysis was carried out in a Jupyter\cite{Basu_undated-vq} analysis environment. 
Joblib\cite{Varoquaux2009-al}, SciPy\cite{Virtanen2020-wt}, Matplotlib\cite{Hunter2007-nq}, 
Seaborn\cite{Waskom2021-gj}, Pandas\cite{McKinney2012-ta} and TQDM\cite{Da_Costa-Luis2019-lr} 
packages were utilized during analysis.

\subsection{Supplementary Information}

All analysis code, data processing code, and supplementary material associated with this manuscript can be found at
 https://github.com/NuttyLogic/MSEPMManuscript. The methylation simulation utility can be found at 
 https://github.com/NuttyLogic/MethSim. The data supporting these findings are openly available at GEO 
 under the series GSE87640, GSE87648, GSE51057, GSE51032, GSE87571, GSE125105, GSE42861, GSE69138, GSE111629, 
 GSE128235, GSE121633, GSE73103, GSE61496, GSE59065, GSE97362, GSE156994, GSE128064 and GSE43976.

\end{linenumbers}

\printbibliography

\begin{center}
    \begin{figure}
    \includegraphics[scale=.25]{Figures/Figure1.png}    
    \footnotesize
    \caption*{\small \textbf{Figure 1:} Simulated factors and the association with simulated methylation values. (A) 
    Age with a non-linear association with methylation (B). Sex (C) with a binary association with methylation (D). 
    Normal factor (E) with a linear relationship with methylation (F). Continuous exponential trait (G) with a linear 
    relationship with methylation. (I) Simulated methylation sites. Each simulation site has a starting methylation 
    value $r^0_i$, rate of change associated with each simulated factor $r_{i, factor}$ and range of 
    variation $\delta_i$. }
    \end{figure}
\end{center}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.2]{Figures/Figure2.png}
    \footnotesize
    \caption*{\small \textbf{Figure 2:} (A) The MAE of the factor predictions on the testing set as multiple factors 
    are modeled simultaneously. (B) Prediction methylation MAE as factors are included in the MSEPM model. (C) 
    Model coefficients for Age, Sex, Continuous Normal and Continuous Exponential factors for models trained 
    $(n=500)$ with all four simulated factors. (D). Simulated and predicted methylation values for all simulated 
    testing sites across all training folds. }
    \end{figure}
\end{center}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.2]{Figures/Figure3.png}
    \footnotesize
    \caption*{\small \textbf{Figure 3:} Distribution of age (A) and (B) sex in aggregate blood dataset. (C) 
    Calculated cell type composition and (D) loading plot of principal components of cell 
    type composition in aggregate blood data set. .}
    \end{figure}
\end{center}

\begin{raggedleft}
\begin{figure}
\includegraphics[scale=.15]{Figures/Figure4.png}
\footnotesize
\caption*{\small \textbf{Figure 4:} MSEPM model trained with age, sex, CT-PC1 and CT-PC2 predictions within testing 
set for methylation associated factors (A) age, (B) sex, (C) CT-PC1 and (D) CT-PC2. (E) Observed and predicted
 methylation values for training set has high concordance $(R^2=0.833, MAE=0.035)$}
\end{figure}
\end{raggedleft}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.15]{Figures/Figure5.png}
    \footnotesize
    \caption*{\small \textbf{Figure5: } (A) Site clustering by standardized model coefficients. Sites clusters show 
    distinct relationships with modeled traits. (B-K) Top five enriched transcription factors for clusters 1 - 10. 
    }
    \end{figure}
\end{center}

\end{document}

### 2023-01-16 DataProcessing/SeriesMatrixMetaExtractor.py

@@ -1,0 +1,57 @@
+from TextIterator import OpenTextFile
+
+
+class SeriesMatrixMetaExtractor:
+
+    def __init__(self, series_matrix_path=None, sample_id=None, platform_id_filter='GPL13534',
+                 phenotype_ids=None, stop_descriptor='!series_matrix_table_begin'):
+        self.series_matrix = OpenTextFile(text_file_path=series_matrix_path, yield_end=stop_descriptor)
+        self.sample_id = sample_id
+        self.phenotype_ids = phenotype_ids
+        self.phenotype_matrix = {}
+        self.platform_filter = platform_id_filter
+        self.series_description = {}
+        self.series_matrix_raw_metadata = {}
+
+    def get_meta_data(self):
+        self.get_matrix_meta()
+        self.experiment_id = self.series_matrix_raw_metadata['!Series_geo_accession'][0].strip('"')
+        self.platform_id = self.series_matrix_raw_metadata['!Series_platform_id'][0].strip('"')
+        if self.platform_id == self.platform_filter:
+            self.data_processing = self.get_descriptive_lines('!Sample_data_processing')
+            self.tissue_source = self.get_descriptive_lines('!Sample_source_name_ch1')
+            self.sample_extraction = self.get_descriptive_lines('!Sample_extract_protocol_ch1')
+            self.sample_ids = self.get_sample_ids
+            self.get_matrix_meta()
+            self.parse_metadata()
+
+    def get_descriptive_lines(self, line_id):
+        return [sample.strip('"') for sample in self.series_matrix_raw_metadata[line_id]]
+
+    @property
+    def get_sample_ids(self):
+        sample_ids = [sample.strip('"') for sample in self.series_matrix_raw_metadata[self.sample_id]]
+        for count, sample_id in enumerate(sample_ids):
+            self.phenotype_matrix[sample_id.strip('"')] = {'experiment_id': self.experiment_id,
+                                                           'platform_id': self.platform_id,
+                                                           'tissue_source': self.tissue_source[count],
+                                                           'data_processing': self.data_processing[count],
+                                                           'sample_extraction': self.sample_extraction[count]}
+        return sample_ids
+
+    def get_matrix_meta(self):
+        for count, line in enumerate(self.series_matrix):
+            if len(line) > 1:
+                if line[0] in self.series_matrix_raw_metadata.keys():
+                    self.series_matrix_raw_metadata[f'{line[0]}___{count}'] = line[1:]
[…]

### 2023-01-16 DataProcessing/minfi_pipeline.R

@@ -1,0 +1,45 @@
+
+
+args <- commandArgs(trailingOnly = TRUE)
+
+exp_id = args[1]
+ref_file= args[2]
+output_dir = args[3]
+
+print(exp_id)
+print(ref_file)
+print(output_dir)
+
+library("FlowSorted.Blood.450k")
+library('IlluminaHumanMethylation450kmanifest')
+library('minfi')
+library('BiocParallel')
+
+options(warn = 0)
+idat_files = as.matrix(read.table(ref_file, header=FALSE, sep='', dec='.'))
+
+rgset = read.metharray(idat_files, verbose=TRUE, force=TRUE)
+rgset.ssNoob = preprocessNoob(rgset, dyeMethod="single")
+grset = mapToGenome(rgset.ssNoob)
+
+# calculate beta matrix 
+b_matrix = getBeta(rgset.ssNoob)
+
+# returns dataframe, median Meth / median unmeth / predicted sex
+sex_QC = minfiQC(rgset.ssNoob, fixOutliers=TRUE)
+
+# returns dataframe
+cell_counts = estimateCellCounts(rgset, processMethod="preprocessNoob")
+
+
+gz1 <- gzfile(paste(output_dir, exp_id, '_methmatrix.gz', sep=''), "w")
+write.csv(b_matrix, gz1)
+close(gz1)
+
+gz2 <- gzfile(paste(output_dir, exp_id, '_qc.gz', sep=''), "w")
+write.csv(sex_QC$qc, gz2)
+close(gz2)
+
+gz3 <- gzfile(paste(output_dir, exp_id, '_cell_counts.gz', sep=''), "w")
+write.csv(cell_counts, gz3)
+close(gz3)

### 2023-01-25 Paper/MSEPM-ManuscriptSupplement.tex
Supplementary Information

\documentclass{article}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{caption}
% \usepackage{psfig}
%\usepackage{epic}
%\usepackage{eepic}
\usepackage{colordvi}
%\usepackage{wrapfig}  
\usepackage{afterpage}
%\usepackage{placeins}
\usepackage{graphics,color}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{placeins}
\usepackage{longtable}
\usepackage{color}      % Need the color package
\usepackage{epsfig}
\usepackage{subfiles}
\usepackage[style=nature]{biblatex}
\usepackage{float}
\usepackage{lineno}

\newcommand{\E}{\mathrm{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\cherry}[1]{\ensuremath{\langle #1 \rangle}}
\newcommand{\eopf}{\framebox[6.5pt]{} \vspace{0.2in}}

%\def\l{{\lambda }}
\def\e{{\varepsilon}}
\def\a{{\alpha }}
\def\S{{\Sigma }}
\def\s{{\sigma }}
\def\M{{\mathcal M}}
\def\X{{\mathcal X }}

\bibliography{MSEPM-Manuscript}

\begin{document}

\large{\bf{The Multi-State Epigenetic Pacemaker enables the 
identification of combinations of factors that influence DNA methylation - Supplementary Information}}\\ \\
\noindent{\small Colin Farrell$^{1,3}$, Kalsuda Lapborisuth$^{1}$, 
                 Sagi Snir$^{2}$,\\ and Matteo Pellegrini$^{1,3}$\\ \\
\noindent{\footnotesize 
$^1$Dept. of Molecular, Cell and Developmental Biology; \\ University of
California, Los Angeles, CA 90095, USA;;\\
\noindent
$^2$Dept. of Evolutionary Biology, University of  Haifa, Israel;\\
$^3$Corresponding Authors; colinpatfarrell@g.ucla.edu, matteop@mcdb.ucla.edu 
}

\begin{center}\rule{0.9\linewidth}{0.5pt}\end{center}
\section{Supplementary Table Descriptions}

\textbf{Supplemental Table 1:} 
MSEPM model parameters for MSEPM blood model trained against age, sex, CT-PC1 and CT-PC2.
\\
\textbf{Supplemental Table 2:}
Sample characteristics for GEO samples used in MSEPM blood model training, validation, and testing. 
\\
\textbf{Supplemental Table 3:}
LOLA transcription factor binding results.
\\
\textbf{Supplemental Table 4:}
Simulated methylation site parameters.
\\
\section{Supplementary Figures}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.2]{Figures/SupplementalFigure1.png}
    \footnotesize
    \caption*{\small \textbf{Supp. Figure 1: } Simulated methylation site intercept accurately modeled with 
    MSEPM four factor model (A).  Simulation phenotype  (B) and site methylation (C) prediction 
    MAE for four-factor MSEPM models fit with a varying number of training samples and sites (D-E).

    }
    \end{figure}
\end{center}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.1]{Figures/SupplementalFigure2.png}
    \footnotesize
    \caption*{\small \textbf{Supp. Figure 2: } MSEPM testing blood model predictions for MSEPM model fit 
    with only age (first row), age / sex (second row), age / sex / cell type PC1 (third row), and age / 
    sex / cell type PC1 / cell type PC2 (fourth row).
    }
    \end{figure}
\end{center}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.1]{Figures/SupplementalFigure3.png}
    \footnotesize
    \caption*{\small \textbf{Supp. Figure 3: } Simulated testing model predictions for MSEPM model fit with only age 
    (first row), age / sex (second row), age / sex / CN (third row), and age / sex / 
    cell type PC1 / cell type PC2 (fourth row).
    }
    \end{figure}
\end{center}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.4]{Figures/SupplementalFigure4.png}
    \footnotesize
    \caption*{\small \textbf{Supp. Figure 4: } Cell type principal component analysis scree plot.

    }
    \end{figure}
\end{center}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.2]{Figures/SupplementalFigure5.png}
    \footnotesize
    \caption*{\small \textbf{Supp. Figure 5: } Pairwise bivariate distributions, and single factor distributions 
    plots, for GEO factor data (age, sex, CT PC1 and CT PC2) used in MSEPM model training. 
    }
    \end{figure}
\end{center}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.18]{Figures/SupplementalFigure6.png}
    \footnotesize
    \caption*{\small \textbf{Supp. Figure 6: } Methylation site and sample factor prediction error for models 
    trained with 1 to 4 factors for training, validation and testing sets. 
    }
    \end{figure}
\end{center}

\begin{center}
    \begin{figure}
    \includegraphics[scale=.1]{Figures/SupplementalFigure7.png}
    \footnotesize
    \caption*{\small \textbf{Supp. Figure 7: } MSEPM blood model predictions for MSEPM model fit against age, sex, 
    CT PC1 and CT PC2 for training, validation and testing sets. 
    }
    \end{figure}
\end{center}

\end{document}

### 2024-05-19 Paper/MSEPM-Manuscript.tex
Manuscript revision

@@
 identification of combinations of factors that influence DNA methylation}}\\ \\
-\noindent{\small Colin Farrell$^{1,3}$, Kalsuda Lapborisuth$^{1}$, 
-                 Sagi Snir$^{2}$,\\ and Matteo Pellegrini$^{1,3}$\\ \\
+\noindent{\small Colin Farrell$^{1,4}$, Keshiv Tandon$^{1}$, Roberto Ferrari$^{2}$, Kalsuda Lapborisuth$^{1}$, 
+                 Sagi Snir$^{3}$,\\ and Matteo Pellegrini$^{1,4}$\\ \\
 \noindent{\footnotesize 
@@
 \noindent
-$^2$Dept. of Evolutionary Biology, University of  Haifa, Israel;\\
-$^3$Corresponding Authors; colinpatfarrell@g.ucla.edu, matteop@mcdb.ucla.edu 
+$^2$Dept. of Chemistry, Life Sciences and Environmental Sustainability, Laboratory of Molecular Cell Biology of the Epigenome (MCBE), University of Parma, Italy;\\
+$^3$Dept. of Evolutionary Biology, University of  Haifa, Israel;\\
+$^4$Corresponding Authors; colinpfarrell@gmail.com, matteop@mcdb.ucla.edu 
 }
@@
 \begin{linenumbers}
-\begin{center}\rule{0.9\linewidth}{0.5pt}\end{center}
 
-Epigenetic clocks, DNA methylation based predictive models of chronological age, are often utilized to study aging 
-associated biology. Despite their widespread use, these methods do not account for other factors that also 
-contribute to the variability of DNA methylation data. For example, many CpG sites show strong sex-specific 
-or cell type specific patterns that likely impact the predictions of epigenetic age.  To overcome these 
-limitations, we developed a multidimensional extension of the Epigenetic Pacemaker, the Multi-State Epigenetic 
-Pacemaker (MSEPM). We show that the MSEPM is capable of accurately modeling multiple methylation associated factors 
-simultaneously, while also providing site specific models that describe the per site relationship between methylation 
-and these factors. We utilized the MSEPM with a large aggregate cohort of blood methylation data to construct models 
-of the effects of age, sex and cell type heterogeneity on DNA methylation.  We found that these models capture a 
-large faction of the variability at thousands of DNA methylation sites.  Moreover, we found modeled sites that are 
-primarily affected by aging and no other factors.  Among these, those that lose methylation over time are enriched 
-for CTCF transcription factor chip peaks, while those that gain methylation over time are enriched for REST 
-transcription factor chip peaks. Both transcription factors are associated with transcriptional maintenance and 
-suggest a general dysregulation of transcription with age that is not impacted by sex or cell type heterogeneity. 
-In conclusion, the MSEPM is capable of accurately modeling multiple methylation associated factors and the models 
-produced can illuminate site specific combinations of factors that affect methylation dynamics.
+\abstract{Epigenetic clocks, DNA methylation based predictive models of chronological age, are often utilized to study aging associated biology. 
+Despite their widespread use, these methods do not account for other factors that also contribute to the variability of DNA methylation data.
+ For example, many CpG sites show strong sex-specific or cell type specific patterns that likely impact the predictions of epigenetic age.  
+ To overcome these limitations, we developed a multidimensional extension of the Epigenetic Pacemaker, the Multi-State Epigenetic Pacemaker (MSEPM). 
+ We show that the MSEPM is capable of accurately modeling multiple methylation associated factors simultaneously, while also providing site specific 
+ models that describe the per site relationship between methylation and these factors. We utilized the MSEPM with a large aggregate cohort of blood
+  methylation data to construct models of the effects of age, sex and cell type heterogeneity on DNA methylation.  We found that these models 
+  capture a large faction of the variability at thousands of DNA methylation sites.  Moreover, this approach allows us to identify sites that 
+  are primarily affected by aging and no other factors.  An analysis of these sites reveals that those that lose methylation over time are 
+  enriched for CTCF transcription factor chip peaks, while those that gain methylation over time are associated with bivalent promoters of 
+  genes that are not expressed in blood. These observation suggest mechanisms that underlie age associated methylation changes and suggest 
+  that age associated increases in methylation may not have strong functional consequences on cell states. In conclusion, the MSEPM is 
+  capable of accurately modeling multiple methylation associated factors and the models produced can illuminate site specific 
+  combinations of factors that affect methylation dynamics.}
 
-\begin{center}\rule{0.9\linewidth}{0.5pt}\end{center}
-\section{Introduction}
+\section{Introduction}\label{sec1}
 
-DNA methylation, the addition of a methyl group to the fifth carbon of the cytosine pyrimidine ring,  is 
-associated with the topological organization of the cellular genome, gene expression and the state of 
-a cell. Within a population of cells the methylation pattern at certain sites can change predictably 
-with the age of the individual from which the cells are drawn. This predictable nature of DNA 
-methylation has led to the development of accurate DNA methylation based predictive models for 
-age and health, termed epigenetic clocks. The difference between the predicted and the expected 
-epigenetic age given an individual's chronological age has been interpreted as a measure of age 
-acceleration\cite{Horvath2018-ia}, and has been associated with mortality\cite{Perna2016-pi,Marioni2015-sn} 
-and other adverse health outcomes\cite{Dugue2018-ad,Huang2019-hf,Armstrong2017-vg,Chuang2017-nk,Horvath2015-af}.
+DNA methylation, the addition of a methyl group to the fifth carbon of the cytosine pyrimidine ring,  is associated with 
+the topological organization of the cellular genome, gene expression and the state of a cell. Within a population of cells 
+the methylation pattern at certain sites can change predictably with the age of the individual from which the cells are 
+drawn. This predictable nature of DNA methylation has led to the development of accurate DNA methylation based predictive 
+models for age and health, termed epigenetic clocks. The difference between the predicted and the expected epigenetic 
+age given an individual’s chronological age has been interpreted as a measure of age acceleration\cite{Horvath2018-ia}, 
+and has been associated with mortality\cite{Perna2016-pi,Marioni2015-sn} and other adverse health 
+outcomes\cite{Dugue2018-ad,Huang2019-hf,Armstrong2017-vg,Chuang2017-nk,Horvath2015-af}.
 
 However, epigenetic clocks suffer from several limitations that limit the interpretability of their 
-predictions and the underlying mechanisms. Epigenetic clocks are generally trained by using penalized regression 
-based methods that attempt to minimize the difference between the predicted and observed value of  age. As a result, 
-as the error between predicted and observed age is decreased, the associations between age acceleration and mortality 
-disappears\cite{Zhang2019-br}. Second generation epigenetic clocks attempt to resolve this issue by fitting a measure 
-of human health, rather than age, and as a result these clocks are generally more sensitive to individual health 
-status \cite{Lu2019-lg,Levine2018-en,Belsky2020-ha}. However, while the response variable is modified in these 
-clocks the method used to fit the clock is largely the same. Epigenetic clocks are generally trained using 
-regularized regression models, where the likelihood is maximized by minimizing the difference between the 
-observed and predicted response variable subject to the elastic net penalty,$\lambda_1$ and $\lambda_2$. 
-Methylation sites that increase model error and are influenced by other relevant factors such as smoking or obesity, 
-may be discarded during model fitting, thus limiting the ability of this approach to account for the effects of these 
+predictions and the underlying mechanisms. Epigenetic clocks are generally trained by using penalized regression based 
+methods that attempt to minimize the difference between the predicted and observed value of  age. As a result, as the 
+error between predicted and observed age is decreased, the associations between age acceleration and mortality 
+disappears\cite{Zhang2019-br}. Second generation epigenetic clocks attempt to resolve this issue by fitting a 
+measure of human health, rather than age, and as a result these clocks are generally more sensitive to individual 
+health status\cite{Lu2019-lg,Levine2018-en,Belsky2020-ha}. However, while the response variable
+is modified in these clocks the method used to fit the clock is largely the same. Epigenetic clocks are generally 
+trained using regularized regression models, where the likelihood is maximized by minimizing the difference between 
+the observed and predicted response variable subject to the elastic net penalty,$\lambda_1$ and $\lambda_2$. Methylation 
+sites that increase model error and are influenced by other relevant factors such as smoking or obesity, may be 
+discarded during model fitting, thus limiting the ability of this approach to account for the effects of these 
 extraneous factors on epigenetic aging.
@@
 As an alternative to penalized regression based methods we previously developed an evolutionary based model for 
-epigenetic dynamics, the Epigenetic Pacemaker (EPM)\cite{Farrell2020-bn,Snir2016-dv}. The EPM attempts to 
-minimize the difference between observed and predicted methylation values amongst a collection of sites through the 
-implementation of a conditional expectation maximization algorithm\cite{Snir2020-tc}. Under the EPM the observed 
-methylation status of a collection of sites is modeled linearly with respect to an input factor of interest, such 
-as age. A hidden epigenetic state, that is related to the initial factor, but not necessarily linearly, is learned 
-through the course of model fitting. The EPM can capture the non-linear relationship between methylation and 
-age\cite{Snir2019-ii} and outputs an interpretable model for each site. However, both the EPM and regression 
-based methods suffer from the same limitation, which is that they are limited to a single trait predicted by, 
-or used to model, observed methylation patterns. In reality, the observed methylation landscape is likely impacted 
-by a variety of factors that act simultaneously to produce the observed methylome of an individual.
+epigenetic dynamics, the Epigenetic Pacemaker (EPM)\cite{Farrell2020-bn,Snir2016-dv}. The EPM 
+attempts to minimize the difference between observed and predicted methylation values amongst a collection of 
+sites through the implementation of a conditional expectation maximization algorithm\cite{Snir2020-tc}. Under the EPM 
+the observed methylation status of a collection of sites is modeled linearly with respect to an input factor of
+ interest, such as age. A hidden epigenetic state, that is related to the initial factor, but not necessarily 
+ linearly, is learned through the course of model fitting. The EPM can capture the non-linear relationship 
+ between methylation and age\cite{Snir2019-ii} and outputs an interpretable model for each site. However,
+  both the EPM and regression based methods suffer from the same limitation, which is that they are limited
+  to a single trait predicted by, or used to model, observed methylation patterns. In reality, the observed 
+  methylation landscape is likely impacted by a variety of factors that act simultaneously to produce the 
+  observed methylome of an individual.
 
-To overcome this limitation, we have developed a multidimensional extension of the EPM, the 
-Multi-State Epigenetic Pacemaker (MSEPM).  We show that the (MSEPM) can accurately model  site specific 
-methylation variation driven by several factors, and given a trained model, accurately predict the values of the 
-factors associated with an individual's observed methylation profile in both simulated methylation datasets and a 
-large aggregate blood tissue methylation dataset. Importantly, as factors that explain the observed methylation 
-profile of an individual are added to the model the ability to model the factors and methylation values improves. 
-Additionally, we show that sites with similar associations to modeled factors cluster together and are enriched 
-for specific transcription factors.
+To overcome this limitation, we have developed a multidimensional extension of the EPM, the Multi-State 
+Epigenetic Pacemaker (MSEPM).  We show that the (MSEPM) can accurately model  site specific methylation 
+variation driven by several factors, and given a trained model, accurately predict the values of the 
+factors associated with an individual’s observed methylation profile in both simulated methylation 
+datasets and a large aggregate blood tissue methylation dataset. Importantly, as factors that explain 
+the observed methylation profile of an individual are added to the model the ability to model the factors 
+and methylation values improves. Additionally, we show that sites with similar associations to modeled factors 
+cluster together and are enriched for specific transcription factors. Therefore, unlike traditional epigenetic 
+clocks, the MSEPM allows us to study mechanisms that may underlie age associated methylation changes.  In our 
+large dataset of blood samples, we find that sites that increase methylation with age are enriched for bivalent 
+promoters, and are proximal to genes that are lowly expressed in blood.  These results suggest that positively 
+age associated sites may not have a significant functional impact on aging traits. The MSEPM is available as a 
+Python package with scikit-learn style syntax under a MIT license at https://github.com/NuttyLogic/MultistateEpigeneticPacemaker.
 
-\section{Results}
+\section{Methods}\label{sec2}
 
@@
 
-The MSEPM attempts to describe the observed methylation status at any single methylation site as a linear combination 
-of methylation associated factors specific to an individual, termed epigenetic factors. Under this model epigenetic 
-factors are related to observable individual factors, such as chronological age, sex and cell types, but may be 
-transformed relative to observable factors. The epigenetic age factor, for example, often has a non-linear 
-relationship with the observed age \cite{Snir2019-ii}. The MSEPM learns the appropriate transformation during model 
-fitting to describe the observed methylation status linearly in terms of the epigenetic age factor 
-(and not linearly with age). Given a site $i$ and individual $j$ the observed methylation status can be 
-modeled as $\hat{m_{i,j}} = r^0_i + \sum^n_{k=1} f_{j,k} r_{i,k}$, where $k$ epigenetic factors, $f_{j,k}$, 
-are weighted by $k$ site specific parameters, $r_{i,k}$, and offset by a sites specific intercept term, $r^0_i$. 
-Site parameters, $r_{i,k}$ and $r^0_i$, are characteristic of the site and shared amongst all individuals while 
-epigenetic factors, $f_{j,k}$, are characteristic of an individual and are the same across all sites for that 
-individual. As a consequence, observed methylation differences between any two individuals are dependent on 
-individual epigenetic factors $f_{j,k}$. In practice, the observed methylation value is also dependent on a 
-normally distributed error term 
-$\epsilon_{i,j}$, $\hat{m_{i,j}} = r^0_i + \sum^n_{k=1} s_{j,n} r_{i,n} + \epsilon_{i,j}$.
+The MSEPM model describes the observed methylation at site $i$ and for individual $j$, $\hat{m_{i,j}}$, 
+as a weighted linear combination of $k$ individual epigenetic factors $p_{k,j}$. 
+$$\hat{m_{i,j}} = r^0_i + \sum^n_{k=1} p_{j,k} r_{i,k}$$ Where $k$ epigenetic factors are weighted by $k$ site specific 
+epigenetic rates of change, $r_{i,k}$, and offset by a sites specific intercept term, $r^0_i$. Site parameters,
+ $r_{i,k}$ and $r^0_i$, 
+are characteristic of the site and shared amongst all individuals while epigenetic factors, $p_{j,k}$, are characteristic 
+of an individual and are the same across all sites for that individual. In practice, 
+the observed methylation value is also dependent on a normally distributed error term 
+$\epsilon_{i,j}$.
 
-Given an input matrix $\hat{M} = [m_{i,j}]$ the objective of the MSEPM is to find the optimal values of 
-$r_{i,k}$ and $f_{k,j}$ that minimize the  residual sum of square (RSS) error, 
-$\epsilon_{i,j}^2 = (\hat{m_{i,j} – r^0_{i} - \sum^n_{n=1} f_{j,k} r_{i,k} })^2$. This is 
-accomplished through the implementation of a conditional expectation maximization algorithm. The 
-maximum likelihood (ML) values of $r_{i,k}$ and $r^0_i$ for each $i$ methylation site can be solved 
-using ordinary least squares (OLS) regression. Given the ML $r_{i,k}$ site coefficients each $k$ epigenetic 
-factor is then updated by fixing the site coefficients and updating $s_{k,j}$ by minimizing the RSS across 
-all $i$ sites using gradient descent, $f_{k,j}^{n+1} = f_{k,j} - \lambda \nabla F(f_{k,j})$, where $\lambda$ 
-is a specified learning rate. The optimization is accomplished by alternating between optimizing $r_{k,i}$ and 
-$f_{k,j}$ until the reduction in sum of the site RSS is below a specified threshold between iterations or a set 
-number of interactions is reached. Importantly, while the ML values of $f_{j,k}$ are by definition linear with 
-the methylation status at any site, the original input factors for $f_{j,k}$ may not be.
+$$\hat{m_{i,j}} = r^0_i + \sum^n_{k=1} p_{j,k} r_{i,k} + \epsilon_{i,j}$$
 
-Provided a trained MSEPM model and an unobserved methylation matrix, epigenetic factors are estimated by 
-calculating each independent OLS for solution all $i$ sites given the $r_{i,k}$ coefficients set for the 
-respective input factor.  These epigenetic factors can then be used to find the expected methylation value using the 
-trained individual site models where $E[m_{i,j}] = r_{i,0} + F_{j} \dot R_{i}$.
+Under this model epigenetic factors are related to observable individual 
+factors $p_{k,j}^0$, such as chronological age, sex and cell types, but may be transformed relative to observable factors. 
+The epigenetic age factor, for example, often has a non-linear relationship with the observed age\cite{Snir2019-ii}. 
+The MSEPM learns the appropriate transformation during model fitting to describe the observed methylation status 
+linearly in terms of the epigenetic age factor, but not linearly with age.
 
-\subsection{Simulated Methylation Associated Factors}
+Given an input matrix $\hat{M} = [m_{i,j}]$ of methylation values for $i$ sites and matched observable epigenetic factors 
+$\hat{P^0} = [p_{j,k}^0]$ for $j$ individuals the objective of the MSEPM is to find the optimal values of 
+$r_{i,k}$ and $p_{j,k}$ that minimize the  residual sum of square (RSS) error,
 
-We implemented a simulation framework that extends the MSEPM model where the methylation status at site $i$ for an 
-individual $j$ is described by a weighted sum of epigenetic factors, 
-$\hat{m_{i,j}} = r^0_i + \sum^n_{l=1} f_{k,j} r_{i,k} + \epsilon_{i,j}$. As shown in our 
-previous work \cite{Snir2019-ii} the association between a methylation associated input factor and methylation 
-status is not necessarily linear. To account for non-linear associations, $f_{k,j}$, have the 
-form $f_{j,k} =  q_{j,k}^{\gamma_{k}}$, where $\gamma_{k}$ is the transformation between a factor of 
-magnitude $q_{n,j}$ and the epigenetic factor. In practice the value of the $f_{k,j}$ is often unknown and the 
-association between methylation status and $f_{k,j}$ is inferred through $q_{n,j}$. We simulated individuals 
-whose methylation is determined by four factors and their associated epigenetic factors: a uniform factor 
-approximating age with a non-linear association with methylation status 
-($q \sim\mathcal{U}(0,100), s_{Age}=q^{0.5}$, Figure 1A-B), a binary trait resembling sex, 
-linearly associated with methylation status ($q \sim B(1,.5), s_{Sex}=q$, Figure 1C-D), a continuous normal 
-(CN) phenotype resembling a cell type with a linear association with methylation 
-status $(q \sim\mathcal{N}(1, 0.1), s_{CN}=q$, Figure 1E-F), and a continuous exponentially (CE) distributed trait 
-resembling obesity with a linear association with methylation status 
-$(q \sim \frac{1}{20}e^{-x/20}, s_{CE}=q, Figure G-H)$. Methylation sites were simulated by first randomly 
-setting the dynamic range of the methylation site, ($ -1< \delta < 1$), a site intercept, $r^0_i$, and the 
-site error, $(\sigma_i \sim \mathcal{U}(0.025, 0.05)$). Simulated methylation sites were associated with a 
-combination of zero, one, or multiple epigenetic factors. Rates for sites associated with multiple factors were 
-set by sampling from a uniform distribution. Site rates were normalized to describe the dynamic range of the 
-simulated methylation site unless the rates for all phenotypes were zero.
+$$\epsilon_{i,j}^2 = (m_{i,j} - r^0_{i} - \sum^n_{k=1} p_{j,k} r_{i,k})^2$$
 
-We simulated 90 methylation sites (Figure 1I). We then evaluated the MSEPM model as follows. 
-We simulated 1000 samples with the four epigenetic factors described above. We then simulated methylation values
- using the simulated site rates. Simulated samples were then split for training ($n=500$) and testing ($n=500$).
-  MSEPM models were then fitted using the values of the input factors, $q_{k,j}$, as the initial guess for
-   $f_{k,j}$. We generated 1000 simulated datasets and fit MSEPM models using four combinations of input 
-   factors (Age, Age-Sex, Age-Sex-CN, Age-Sex-CN-CE). Within each simulation, epigenetic state predictions and 
-   methylation site predictions were made for all testing samples. All models captured the nonlinear association 
-   between simulated age and methylation (S. Figure 1). As the number of factors in the model is increased the mean 
-   absolute error (MAE) between the predicted epigenetic states and the simulated epigenetic factors decreases 
-   (Figure 2A). Importantly, to accurately assess simulated age it is necessary to account for the influence of 
-   the other simulated factors (Sex, CN, CE). The MAE between the predicted and simulated methylation values decreases 
-   as simulated factors are added to the model, and accurately assessing the methylation status of a simulated site 
-   requires that the factor associated with the methylation status at the site is included in the model (Figure 2A).
+This is accomplished through the implementation of a conditional expectation maximization algorithm. The maximum likelihood (ML) values of 
+$r_{i,k}$ and $r^0_i$ can
+be solved using ordinary least squares (OLS) regression. Provided the ML estimates for $r_{i,k}$, the site coefficients are fixed and 
+epigenetic factors, $p_{j,k}$, are  updated by minimizing the RSS across all $i$ sites using 
+gradient descent, $$p_{j,k}^{n+1} = p_{j,k}^n - \lambda \nabla F(p_{j,k})$$ where $\lambda$ is a specified learning rate. 
+The optimization is accomplished by alternating between optimizing $r_{i,K}$ 
+and $p_{j,k}$ until the reduction in sum of the site RSS is below a specified threshold or a set number 
+of iterations is reached. Importantly, while the ML values of $p_{j,k}$ are 
+by definition linear with the methylation status at any site, the original input factors for $p_{j,k}^0$ may not be.
 
-The MSEPM model generated using all four simulated factors can capture the relative magnitude of the simulated 
-site-specific rates (Figure 2C-F).  However, the model has difficulty capturing the exact relationship between the 
-simulated factors (age, CN and CE) and the inferred factors (Figure 2C, E-F). This is likely due to limitations of 
-the model at capturing nonlinear methylation association and a limited training range for normally and exponentially 
-distributed traits. Regardless, the four-factor model can accurately predict the simulated methylation value 
-(Figure 2 D) and site intercept (Supp. Figure 1G). We also assessed the model robustness to variation in the 
-number of samples and sites used for model training by randomly selecting a reduced subset of samples or sites for 
-model training. MSEPM models trained with age, sex, CN, and CE can accurately assess all simulated phenotypes with 
-few samples and sites (Supp Figure2 B-E).
+Provided a trained MSEPM model and an unobserved methylation matrix, epigenetic factors are estimated by calculating each independent OLS for solution all $i$ 
+sites given the $r_{i,k}$ coefficients set for the respective input factor.  These epigenetic factors can then be used to find the expected methylation value 
+using the trained individual site models where $$E[m_{i,j}] = r_{i,0} + P_{j} \dot R_{i}$$
+
+\subsection{MSEPM Simulation Framework}
+
+We implemented a simulation framework using the MSEPM formulation evaluate the performance of the MSEPM model under various conditions.
+To simulate the association between methylation status and an observable factor we modeled the 
+epigenetic factor $p_{k,j}$ as function of time, or age, and magnitude, $p_{k,j}^0$ with a non-linear transformation $\gamma_{k}$, 
+where $p_{k,j} = Age_j p_{k,j}^{0,\gamma_{k}}$. In practice the value of the $p_{k,j}$ is often unknown and the association between 
+methylation status and $p_{k,j}$ is inferred through the observable factor $p_{k,j}^0$.
+
+Methylation sites were simulated by first randomly setting the dynamic range of the methylation site, 
+$-1< \delta < 1$ a site intercept, $r^0_i$, and the site error, $\sigma_i \sim \mathcal{U}(0.025, 0.05)$. The dynamic 
+range of the methylation site is described by the initial methylation value, $m_0 ~ \beta(.2,.2)$, 
+and the target methylation value, $m_t$, where the dynamic range is $\delta_i = m_t - m_o$.
+The value of $m_t$ is set conditionally to ensure dynamic range is always larger than some specified threshold, $\theta$, where 
+$ \theta \leq|\delta| \geq. r^0_i ~ \beta(.2,.2)$.
+
+Simulated methylation sites are then randomly associated with a combination of zero, one, or multiple epigenetic factors. Rates for sites 
+associated with multiple factors were set by sampling from a uniform distribution. The weighted factor rates are normalized so 
+the input combination of traits describes the dynamic range of the simulated site, $\delta$. If a site is associated 
+with no factors the observed methylation status of a site is described by a random normal with a characteristic offset, $\hat{m_i} = r_{0,i} + N(\mu, \sigma)$.
+
+\subsection{Blood MSEPM Model Training}
+MSEPM models were trained using a large aggregate dataset of blood derived methylation data from 17 publicly available 
+datasets\cite{Ventham2016-qj,Demetriou2013-wb,Polidoro2013-xv,Johansson2013-of,Arloth2020-lo,Liu2013-dg,Soriano-Tarraga2016-uq,Chuang2017-nk,
+Zannas2019-me,Kurushima2019-pe,Voisin2015-lh,Tan2014-sg,Tserel2015-ro,Butcher2017-oz,Dabin2020-iw,Marabita2013-cj,ValleG2019-xi}. 
+Illumina methylation 450K Beadchip methylation array IDAT files were processed using minfi\cite{Aryee2014-ky} (v1.34.0). 
+Sample IDAT files were processed in batches according to GEO series and Beadchip  identification. Methylation values within 
+each batch were normal-exponential normalized using out-of-band probes\cite{Triche2013-pp}. Blood cell types counts were estimated 
+using a regression calibration approach\cite{Houseman2012-rr} and sex predictions were made using the median intensity measurements
+of the X and Y chromosomes as implemented in minfi\cite{Aryee2014-ky}. Samples were filtered for quality control using the the relative 
+intensity of the methylated and unmethylated probes. Samples were used for downstream analysis if the sample median 
+methylation probe intensity was greater than 10.5 and the difference between the observed and expected median 
+unmethylation probe intensity is less than 0.4, where the expected median unmethylated intensity is described by 
+$E[intensity_{unmethylated}] = 0.66 intensity_{methylated} + 3.718$. This resulted in a total of 5687 samples.
+
+We trained MSEPM models using data assembled from four GEO series\cite{Johansson2013-of,Liu2013-dg,Butcher2017-oz,Damaso2020-gd}  ($n=1605$). 
+The samples were randomly split into training ($n=1203$) and validation ($n=402$) sets stratified by age. Methylation values for all 
+samples were quantile normalized by probe type\cite{Horvath2013-sk} using the median site methylation values across all training samples for each methylation site.
+Training set blood cell type abundance estimates were used to train a principal component analysis (PCA) model which was then used to calculate cell 
+type PCA estimates for the validation and testing sets. Methylation sites were selected for modeling with MSEPM if the site methylation values were correlated with age ($n=276$) , 
+sex ($n=49$), CT-PC1 ($n=120$), CT-PC2 ($n=116$) or a combination of factors ($n=238$) by absolute pearson correlation coefficient. Where a absolute pearson correlation coefficient greater than 0
+.7, 0.995, 0.92 and 0.64 for age, sex, CT-PC1 and CT-PC2 respectively. Sites with a sum of absolute pearson coefficients across the four factors greater than 1.8 were also included $(n=238)$ 
+for a total of 778 methylation sites.
+Min-max, (0-1), scalers were fit using the training input features.
+Validation and testing sample features were transformed with the trained scalers.
+Age was min-max
+scaled on a range from 0-100 years. MSEPM models were trained with a learning rate of 0.01 with an iteration limit of 200.
+
+\subsection{Blood MSEPM Model Cluster Transcription Factor Overlap Analysis}
+We evaluated the relationship between modeled sites, input factors and regulatory transcription factors using overlap enrichment analysis.  
+We built a custom transcription factor reference set using ENCODE V4 transcription factor chromatin immunoprecipitation\cite{ENCODE_Project_Consortium2012-oe,Davis2018-ha} 
+(release 1.4.0 - 2.1.2)  irreproducible discovery rate narrow bed peaks, 
+  which contains peaks with high rank consistency between replicates, that were not audited for non-compliance or errors. 
+  GRCh38 region coordinates were lifted to GRCh37 coordinates using liftOver\cite{Hinrichs2006-oq}. The overlap reference 
+  contains 714 transcription factor targets from 1621 accession IDs.
+
+We then performed hierarchical clustering of the four factor MSEPM model sites based on the similarity of their regression coefficients.
+Individual methylation site coefficients were first normalized by the standard deviation of methylation values of the site among the training samples, $r_{i,k} / \sigma_{i}$. 
+A distance matrix was then created by taking the Euclidean distance between the normalized site model coefficients. Sites were then clustered using Ward’s 
+method which seeks to minimize within cluster variance by minimizing the increase in the error sum of squares (ESS) through successive cluster fusions.  
+Clusters label by tree cutting at a height of 18. All clustering analysis was carried out using SciPy v1.6.3\cite{Virtanen2020-mm}.
+
+Transcription factor enrichment analysis was performed with  LOLA\cite{Sheffield2016-wg} which assesses the genomic 
+region set overlap between a set of query regions and a set of reference regions, within a specified shared background set, 
+using Fisher's exact test. Overlap analysis was performed for sites within a cluster against the ENCODE V4 reference region 
+(1BP minimum overlap) using all sites assayed with Infinium HumanMethylation450K BeadChip as background.
+
+\subsection{Clustering sites with age-associated increases in methylation}
+To better understand age associated methylation in whole blood, we examined each site within MSEPM four factor blood model cluster 7 individually, 
+as this cluster contains sites that have methylation that increases with age but is not strongly affected by other factors. 
+Using the EWAS Data Hub (Xiong, et al. 2016), we validated our results by obtaining additional methylation by age data in 
+whole blood for each site in the cluster (McCartney, et al. 2019).  We created a matrix with every sample and its associated 
+methylation and age from cluster 7, then used age associated methylation levels to create a clustered heatmap using the Matlab 
+function Clustergram. We then clustered the tree into four groups which were analyzed separately.
+
+We also identified the genes that were proximal to each site using Cistrome-GO (Li et al. 2019). We then 
+examined the expression of the genes across tissues in the Genotype-Tissue Expression (GTEx) database database. 
+We used the GTEx Multi Gene Query to find which tissues those genes belonged to.
+
+We utilized the Toolkit for Cistrome Data Browser \cite{Zheng2019-vm,Mei2017-yq} for the analysis of significant factors in each cluster. This allowed 
+us to input .bed files of each sub-cluster and generate a GIGGLE score for specific transcription factors, histone marks, 
+and chromatin regions to assess significance of these elements. A GIGGLE score tailored ranking of loci based on overlap 
+of genomic features provided by the user\cite{Layer2018-gr}.
+
+\subsection{H3K4me3 enrichment analysis} 
+Enrichment of analysis for H3K4me3 (figure 7A) was carried out by downloading rpm normalized bigwig files of H3K4me3 ChIP-seq data from 
+epigenomesportal\cite{Bujold2016-vk} for CD38+ B Cells and CD56+ NTK Cells (for both 0-5 years old and 60-65 years old individuals). 
+Heatmaps of H3K4me3 were generated using deepTools2\cite{Ramirez2016-xl} using the computeMatrix and plotHeatmap function to plot the 
+bigwig signal over genomic regions of cluster 7 as the BED input. The IGV genome browser\cite{Robinson2011-he} was used to generate an image of the KCTD1 
+and IRS2 promoter regions shown in figure 7B using downloaded bigwig tracks.
+
+\subsection{Analysis Environment}
+Analysis was carried out in a Jupyter\cite{Basu_undated-vq} analysis environment. 
+Joblib\cite{Varoquaux2009-al}, SciPy\cite{Virtanen2020-wt}, Matplotlib\cite{Hunter2007-nq}, 
+Seaborn\cite{Waskom2021-gj}, Pandas\cite{McKinney2012-ta} and TQDM\cite{Da_Costa-Luis2019-lr} 
+packages were utilized during analysis.
+
+\section{Results}\label{sec3}
+
+\subsection{Simulated Methylation Associated Traits}
+
+We simulated individuals whose methylation is determined by four factors and their associated epigenetic factors: 
+a uniform factor approximating age with a non-linear association with methylation status 
+$$q \sim\mathcal{U}(0,100), s_{Age}=q^{0.5}, \text{Figure 1A-B}$$ a binary trait resembling sex, linearly associated 
+with methylation status $$q \sim B(1,.5), s_{Sex}=q, \text{Figure 1C-D}$$ a continuous normal (CN) phenotype 
+resembling a cell type with a linear association with methylation status $$q \sim\mathcal{N}(1, 0.1), s_{CN}=q, 
+\text{Figure 1E-F}$$ and a 
+continuous exponentially (CE) distributed trait resembling obesity with a linear association with methylation status 
+$$q \sim \frac{1}{20}e^{-x/20}, s_{CE}=q, \text{Figure 1G-H})$$
+
+We simulated 90 methylation sites (Figure 1I). We then evaluated the MSEPM model as follows. We simulated 1000 samples with the 
+four epigenetic factors described above. We then simulated methylation values using the simulated site rates. Simulated samples
+ were then split for training ($n=500$) and testing ($n=500$). MSEPM models were then fitted using the values of the input factors, $p^0_{k,j}$. 
+ We generated 1000 simulated datasets and fit MSEPM models using four combinations of input factors (Age, Age-Sex, Age-Sex-CN, Age-Sex-CN-CE). 
+ Within each simulation, epigenetic state predictions and methylation site predictions were made for all testing samples. All models captured 
+ the nonlinear association between simulated age and methylation (Supp. Figure 1). As the number of factors in the model is increased the mean 
+ absolute error (MAE) between the predicted epigenetic states and the simulated epigenetic factors decreases (Figure 2A). Importantly, to accurately
+  assess simulated age it is necessary to account for the influence of the other simulated factors (Sex, CN, CE). The MAE between the predicted 
+  and simulated methylation values decreases as simulated factors are added to the model, and accurately assessing the methylation status of a 
+  simulated site requires that the factor associated with the methylation status at the site is included in the model (Figure 2A).
+
+The MSEPM model generated using all four simulated factors can capture the relative magnitude of the simulated site-specific rates (Figure 2C-F).  
+However, the model has difficulty capturing the exact relationship between the simulated factors (age, CN and CE) and the inferred 
+factors (Figure 2C, E-F). This is likely due to limitations of the model at capturing nonlinear methylation association and a 
+limited training range for normally and exponentially distributed traits. Regardless, the four-factor model can accurately 
+predict the simulated methylation value (Figure 2 D) and site intercept (Supp. Figure 1A). We also assessed the model 
+robustness to variation in the number of samples and sites used for model training by randomly selecting a reduced 
+subset of samples or sites for model training. MSEPM models trained with age, sex, CN, and CE can accurately assess 
+all simulated phenotypes with few samples and sites (Supp. Figure2 B-E).
 
 \subsection{Blood MSEPM Model}
+We next applied the MSEPM to real data.
+We utilized a large aggregated dataset composed of Illumina 450k array data from 17 publicly available datasets\cite{Ventham2016-qj,Demetriou2013-wb,Polidoro2013-xv,Johansson2013-of,Arloth2020-lo,Liu2013-dg,Soriano-Tarraga2016-uq,Chuang2017-nk,
+Zannas2019-me,Kurushima2019-pe,Voisin2015-lh,Tan2014-sg,Tserel2015-ro,Butcher2017-oz,Dabin2020-iw,Marabita2013-cj,ValleG2019-xi}
+ deposited in the Gene Expression Omnibus\cite{Barrett2012-gu} (GEO) generated from blood derived samples (whole blood, peripheral blood lymphocytes, and 
+peripheral blood mononuclear cells). The aggregate data spanned a wide age range (0.0  - 99.0 years, Figure 3A), contained more predicted females ($n=3392$) than males ($n=2295$, Figure 3B) 
+and reasonable predicted cell type abundance estimates (Figure 3C). The first principal component of a PCA modle trained cell type abundance estimates (CT-PC1) is 
+largely driven by the relative abundance of granulocytes (Figure 3D), while the second PC (CT-PC2) captures relative 
+differences in the abundance of differentiated lymphocytes (Figure 3D).
 
-We utilized a large aggregated dataset composed of Illumina 450k array data from 17 publicly available 
-datasets \cite{Ventham2016-qj,Demetriou2013-wb,Polidoro2013-xv,Johansson2013-of,Arloth2020-lo,Liu2013-dg,Soriano-Tarraga2016-uq,Chuang2017-nk,Zannas2019-me,Kurushima2019-pe,Voisin2015-lh,Tan2014-sg,Tserel2015-ro,Butcher2017-oz,Dabin2020-iw,Marabita2013-cj,ValleG2019-xi}
-deposited in the Gene Expression Omnibus \cite{Barrett2012-gu} (GEO) generated from blood derived samples 
-(whole blood, peripheral blood lymphocytes, and peripheral blood mononuclear cells). All datasets were processed 
-through a unified pipeline from raw array intensity data (IDAT) files using minfi (Aryee et al., 2014). Sex and 
-blood cell type abundance predictions were made for each as previously described\cite{Houseman2012-rr,Aryee2014-ky}. 
-The aggregate data spanned a wide age range (0.0  - 99.0 years, Figure 3A), contained more predicted females 
-($n=3392$) than males ($n=2295$, Figure 3B) and produced reasonable cell type abundance estimates (Figure 3C).
-
-We trained MSEPM models using data assembled from four GEO 
-series\cite{Johansson2013-of,Liu2013-dg,Butcher2017-oz,Damaso2020-gd}  ($n=1605$) with samples spanning a wide age 
-range (0.01 - 94.0 years). The samples were randomly split into training ($n=1203$) and validation ($n=402$) sets 
-stratified by age. Training set blood cell type abundance estimates were used to train a principal component analysis 
-(PCA) model which was then used to calculate cell type PCA estimates for the validation and testing sets. The first 
-cell type principal component (CT-PC1) is largely driven by the relative abundance of granulocytes (Figure 3D), while 
-the second PC (CT-PC2) captures relative differences in the abundance of differentiated lymphocytes (Figure 3D).  
-Methylation values for all samples were quantile normalized by probe type \cite{Horvath2013-sk} using the median 
-site methylation values across all training samples for each methylation site.
-
-Methylation sites were selected for modeling with MSEPM if the sites were highly correlated with 
-age ($n=276$) , sex ($n=49$), CT-PC1 ($n=120$), CT-PC2 ($n=116$) or a combination of factors ($n=238$), 
-a total of 778 unique sites. Four MSEPM models were fit using the selected sites with four combinations of input 
-factors (Age, Age Sex, Age Sex CT-PC1, and Age Sex CT-PC1 CT-PC2). Training sample factors were min-max scaled between 
-0 and 1 before model training. The association between the fit epigenetic factor predictions against the input modeled 
-factors was assessed by fitting a trendline between epigenetic state predictions and scaled continuous input factors 
-using the state prediction made for the MSEPM model trained with all four input factors. Performance of the MSEPM 
-model was then evaluated using the testing samples ($n=4,082$). The performance of the MSEPM largely closely 
-resembles the simulation results. All four MSEPM models capture the nonlinear relationship between age and 
-methylation status (Supp. Figure 6). The epigenetic state prediction associated with age improves as the underlying 
-methylation data are more fully explained through the addition of epigenetic factors (Supp. Figure 6). The MSEPM model 
-fit with Age, Sex, CT-PC1 and CT-PC2 can accurately model the associated epigenetic state for each factor (Figure 4 A-D) 
-and accurately predicts the methylation levels at individual sites ($R^2=0.935$, $MAE=0.035$, Figure 4 E). 
-The trained MSEPM produces a collection of methylation site models that can help explain the association between 
-modeled factors and methylation status.
+We trained MSEPM models using methylation sites ($n=778$) that were correlated with the observable input factors. 
+MSEPM models were fit using four combinations of input factors (Age, Age Sex, Age Sex CT-PC1, and Age Sex CT-PC1 CT-PC2). 
+The association between the fit epigenetic factor predictions against the input modeled factors was assessed by 
+fitting a trendline between epigenetic state predictions and scaled continuous input factors using the state 
+prediction made for the MSEPM model trained with all four input factors. Performance of the MSEPM model was then 
+evaluated using the testing samples ($n=4,082$). The performance of the MSEPM largely closely resembles the 
+simulation results. All four MSEPM models capture the nonlinear relationship between age and 
+methylation status (Supp. Figure 6). The epigenetic state prediction associated with age improves as 
+the underlying methylation data are more fully explained through the addition of epigenetic factors (Supp. Figure 6).
+The MSEPM model fit with Age, Sex, CT-PC1 and CT-PC2 can accurately model the associated epigenetic state 
+for each factor (Figure 4 A-D) and accurately predicts the methylation levels at individual sites ($R^2=0.935$, $MAE=0.035$, Figure 4 E). 
+The trained MSEPM produces a collection of methylation site models that can help explain the association between modeled factors and methylation status.
 
 \subsection{Analysis of chromatin regulators of site clusters}
+We evaluated the relationship between sites that are influenced by age, sex, CT-PC1 or CT-PC2  and potential regulatory 
+factors by performing overlap enrichment analysis of these sites with transcription factor chromatin immunoprecipitation 
+peaks present in the ENCODE V4\cite{Davis2018-ha,ENCODE_Project_Consortium2012-oe} release. 
+We first identified sites with similar coefficients of epigenetic factors through hierarchical clustering. 
+The resulting tree was cut at a height of 18 to produce 10 distinct clusters with clear associations to the modeled factors (Figure 5A).
 
-We evaluated the relationship between sites that are influenced by age, sex, CT-PC1 or CT-PC2  and potential 
-regulatory factors by performing overlap enrichment analysis of these sites with transcription factor chromatin 
-immunoprecipitation peaks present in the ENCODE V4 \cite{ENCODE_Project_Consortium2012-oe,Davis2018-ha} release. We 
-first identified sites with similar coefficients of epigenetic factors through hierarchical clustering. Briefly, 
-sites were clustered using Ward's method by the euclidean distance between site regression coefficients that were 
-normalized by the standard deviation of the methylation values amongst the training samples. The resulting tree was 
-cut at a height of 18 to produce 10 distinct clusters with clear associations to the modeled factors (Figure 5A).
+The site clusters largely conform to underlying biological expectations. Cluster one contains sites 
+that are wholly associated with sex status and localized to the X chromosome (Supp. Table 1) and is enriched for peaks of 
+transcription factors associated with sex specific regulation such as MAZ\cite{Lopes-Ramos2020-ex}. 
+Clusters nine and ten contain sites whose methylation status is largely driven by CT-PC1, and are enriched for 
+transcription factors associated with granulocyte development (CEBPB, CEBPA, EP300, ETV6)\cite{Theilgaard-Monch2022-zw,Guerzoni2006-ii}. 
+Similarly, clusters two, five and eight are associated with CT-PC2 and are enriched for 
+transcription factor peaks associated with immune development (ZBED1, ETV6, FOSL2, FOS, TBX21). Clusters four and 
+six are associated with loss of methylation with age. Cluster six is highly enriched for CTCF binding sites; CTCF 
+is known to increase at sites where methylation is lost during aging\cite{Tharakan2020-pj}. Cluster four is enriched 
+for STAT3 whose activation during exercise is age dependent\cite{Mohamed2020-he,Trenerry2008-kj}.  
+Cluster seven is associated with the accumulation of methylation with age and is enriched for immunoprecipitation peaks
+ for aging associated transcription factors SMAD4 and RE1-Silencing Transcription Factor (REST). SMAD4 encodes a protein 
+ involved in the transforming growth factor beta (TGF-$\beta$) signaling pathway. Age related dysregulation of TGF-$\beta$ 
+ has been linked to reduced skeletal muscle regeneration\cite{Carlson2009-uz,Paris2016-fo} and SMAD4 polymorphisms 
+ are associated with longevity\cite{Carrieri2004-by}). REST is a transcriptional repressor of neuron specific genes in 
+ non-neuronal cells\cite{Chong1995-dj,Coulson2005-pb}. REST expression is upregulated in aged prefrontal cortex 
+ tissue and the absence of REST expression is associated with cognitive impairment\cite{Lu2014-dz} and 
+ cellular senescence in neurons\cite{Rocchi2021-od}).
 
-The site clusters largely conform to underlying biological expectations. Cluster one contains sites that are wholly 
-associated with sex status and localized to the X chromosome (Supp. Table 1) and is enriched for peaks of 
-transcription factors associated with sex specific regulation such as MAZ \cite{Lopes-Ramos2020-ex}. Clusters nine and 
-ten contain sites whose methylation status is largely driven by CT-PC1, and are enriched for transcription factors 
-associated with granulocyte development (CEBPB, CEBPA, EP300, ETV6)\cite{Theilgaard-Monch2022-zw} \cite{Guerzoni2006-ii}.  
-Similarly, clusters two, five and eight are associated with CT-PC2 and are enriched for transcription factor peaks 
-associated with immune development (ZBED1, ETV6, FOSL2, FOS, TBX21). Clusters four and six are associated with loss of 
-methylation with age. Cluster six is highly enriched for CTCF binding sites; CTCF is known to increase at sites where 
-methylation is lost during aging \cite{Tharakan2020-pj}. Cluster four is enriched for STAT3 whose activation during 
-exercise is age dependent \cite{Trenerry2008-kj,Mohamed2020-he}.  Cluster seven is associated with the accumulation 
-of methylation with age and is enriched for immunoprecipitation peaks for aging associated transcription factors 
-SMAD4 and RE1-Silencing Transcription Factor (REST). SMAD4 encodes a protein involved in the transforming growth 
-factor beta (TGF-$\beta$) signaling pathway. Age related dysregulation of TGF-$\beta$ has been linked to reduced skeletal muscle 
-regeneration\cite{Paris2016-fo,Carlson2009-uz} and SMAD4 polymorphisms are associated 
-with longevity \cite{Carrieri2004-by}. REST is a transcriptional repressor of neuron specific genes in non-neuronal 
-cells \cite{Chong1995-dj,Coulson2005-pb}. REST expression is upregulated in aged prefrontal cortex tissue and the 
-absence of REST expression is associated with cognitive impairment \cite{Lu2014-dz} and 
-cellular senescence in neurons \cite{Rocchi2021-od}.
+\subsection{Analysis of sites with age-associated increases in methylation}
+Because of our interest in the mechanisms that underlie ages associated increase in methylation, we focused on cluster seven,  
+as these sites have methylation increases that depend primarily on age rather than sex and cell types. Cluster 7 consisted 
+of 93 CpG sites. To obtain an independent measure of how these sites change with age, we obtained age associated methylation
 
-\section{Discussion}
+data from the EWAS Data Hub\cite{Xiong2020-fa}, with a focus on whole blood methylation. The dataset consisted of  
+about 1600 individuals with ages ranging from 0 to 113 years old\cite{McCartney2019-qi}. We  clustered the sites 
+based on age associated methylation levels, meaning the rate of methylation based on age for each marker. Each site 
+was organized into an ordered matrix with methylation levels at each age, then grouped into four sub-clusters: A, B, C, and D.
+ As seen in Figure 6A, Cluster A had the highest average methylation across ages, and each consecutive cluster had a 
+ decrease in average methylation. We next examined chromatin accessibility, transcription factors, histone marks, and 
+ genes associated with each cluster.  As shown in Supp. Figure 7, genes proximal to Cluster 7 sites were lowly expressed 
+ in blood compared to other tissues.  We analyzed chromatin accessibility, transcription factors, and histone marks 
+ associated with these four groups. We computed levels of H3K27ac, H3K27me3, H3K4me3, and H3K9me3 across the four 
+ subclusters. As seen in Figure 6C, H3K4me3 increased from clusters A through D. Figure 6E shows that H3K27ac
+  increased from clusters A through C, but then decreased in D. These results suggest that subcluster D is 
+  enriched for bivalent domains, characterized by H3K4me3 and H3K27me3.
 
-Epigenetic clocks are widely used tools to study human aging and health. Despite their widespread use, the 
-biological interpretation of model parameters is limited. A methylome is influenced by many different biological 
+Based on these results we hypothesize that the mechanisms that underlies the gain of methylation with age at these 
+bivalent promoters is the age-associated loss of H3K4me3.  It is well established that the presence of trimethylation
+ on H3K4 inhibits de novo methylation, and this effect explains the hypomethylation that is typical of promoters, 
+ including bivalent promoters.  We therefore hypothesize that the gain of methylation at these sites may be caused 
+ by an age associated loss of H3K4me3.  In order to demonstrate that H3K4me3 decreases with age for genomic regions 
+ where DNA methylation increases, we used published H3K4me3 ChIP-seq data from epigenomesportal\cite{Bujold2016-vk}. 
+ We selected two different blood cell types CD38+ B Cells and CD56+ NTK Cells and plotted the H3K4me3 signal of young 
+ (0 to 5 years old) versus old individuals (60 to 65 years old) over genomic regions of cluster 7 (Figure 7A). 
+ Our analysis shows that younger individuals have higher levels of H3K4me3 compared to older ones (Figure 7A) as 
+ also shown for two selected genomic loci of cluster 7 (the promoters of KCTD1 and IRS2 genes) where we can observe
+ a marked decrease in the levels of H3K4me3 as age increases (Figure 7B). All together these data suggest that genomic 
+ regions whose DNA methylation is increased with age exhibit an age dependent loss of H3K4me3, thus showing an inverse 
+ correlation between DNA methylation and H3K4me3 at these genomic loci.
+
+\section{Discussion}\label{sec4}
+
+Epigenetic clocks are widely used tools to study human aging and health. Despite their widespread use, 
+the biological interpretability of the models is limited. A methylome is influenced by many different biological 
 processes occurring simultaneously over time that may differ among individuals. Epigenetic clocks, while producing
@@
   the penalized regression based methods used to fit most epigenetic clocks select sites that minimize, or regress out,
-the influence of other factors and omit groups of sites that are correlated. To overcome these limitations, here we 
-propose a multidimensional extension of the EPM model, the MSEPM.
+   the influence of other factors and omit groups of sites that are correlated. To overcome these limitations, here 
+   we propose a multidimensional extension of the EPM model, the MSEPM.
 
-In contrast to previous methods, the MSEPM aims to simultaneously model the effect of multiple factors on the 
-methylome.The simulation and blood MSEPM models show that concurrently modeling age, cell type composition and sex 
-can minimize  model residuals when compared with the MSEPM model fit with age only. The residual of the age only 
-model is often interpreted as a measure of age acceleration. When multiple methylome associated traits are modeled 
-simultaneously this residual can be explained directly by other factors and the association between the methylome 
-and a trait of interest can be inferred.
+In contrast to previous methods, the MSEPM aims to simultaneously model the effect of multiple factors on the methylome.
+The simulation and blood MSEPM models show that concurrently modeling age, cell type composition and sex can minimize  
+model residuals when compared with the MSEPM model fit with age only. The residual of the age only model is often 
+interpreted as a measure of age acceleration. When multiple methylome associated traits are modeled simultaneously 
+this residual can be explained directly by other factors and the association between the methylome and a trait of 
+interest can be inferred.
 
@@
 the blood MSEPM model conforms to expected biology. Sites with a strong sex association localize to the X chromosome
-and sites associated with cell types are enriched for transcription factors associated with the development of 
-immune cells. Sites that are primarily affected only by age in the blood MSEPM model are of particular interest. 
-As others have previously described, sites that progressively lose methylation over time are strongly enriched for 
-CTCF \cite{Han2020-zj,De_Lima_Camillo2022-lu}, while in the blood MSEPM model sites that gain methylation over time 
-are enriched for REST. The association between age and REST is well established, but the association between gain of 
-methylation and age has previously not been reported. Both transcription factors are associated with the maintenance 
-of transcriptional profiles and suggest that there is a general dysregulation of maintenance of cell specific 
-chromosomal organization over time.
+ and sites associated with cell types are enriched for transcription factors associated with the development of immune cells.
 
-In conclusion, we introduced a multi-dimensional extension of the Epigenetic Pacemaker, the MSEPM. The MSEPM is 
-capable of accurately modeling multiple methylation associated factors simultaneously. This paradigm can elucidate 
-the site specific regulation underpinning methylome dynamics. The MSEPM is available under the MIT license at
- https://github.com/NuttyLogic/MultistateEpigeneticPacemaker.
+CpG sites that are primarily affected only by age in the blood MSEPM model are of particular interest. As 
+others have previously described, sites that progressively lose methylation over time are strongly enriched for
+ CTCF\cite{De_Lima_Camillo2022-lu,Han2020-zj}.  As CTCF plays a key role in long range chromatin interactions, 
+ this may suggest that there are age-associated changes in three dimensional chromatin structure, and that the  
+ structure may become more disordered with age. In fact, alterations in CTCF binding and function with age 
+ have been implicated in the pathogenesis of various age-related diseases, including cancer. For example, 
+ changes in the chromatin structure and gene expression due to altered CTCF binding can contribute to the 
+ genomic instability and altered cell proliferation characteristic of cancerous cells (Hnisz et al., 2016; Phillips et al., 2009).
 
-\section{Methods}
+We identified a cluster of sites that showed increasing methylation with age and that were not significantly 
+affected by other factors.  We found that these sites are enriched for the transcription factor REST. The RE1-Silencing 
+Transcription Factor (REST), also known as Neuron-Restrictive Silencer Factor (NRSF), is a key regulatory protein 
+involved in the development and differentiation of neurons. It plays a crucial role in neurogenesis, neuronal 
+differentiation, and in the maintenance of the neuronal phenotype by regulating gene expression\cite{Schoenherr1995-fc}. 
+REST achieves this by binding to the neuron-restrictive silencer element (NRSE) or RE1 sites in the DNA, leading to the 
+repression of gene transcription in non-neuronal cells or in neuronal progenitor cells, ensuring that neuronal genes 
+are expressed only in neurons\cite{Chong1995-dj,Ooi2007-kk,Bruce2004-je}.  The fact that this factor 
+is enriched at the positively age-associated sites suggests that these sites are likely expressed in neuronal cells 
+but not in blood.  In fact this is what we find when we examine the tissue specific expression of the genes proximal 
+o these sites.
 
-\subsection{MSEPM}
+We also examined the histone modifications associated with the positively age-associated sites and found that they were 
+enriched for H3K4me3 and H3K27me3.  These sites are characteristic of bivalent promoters.  Bivalent promoters play a 
+crucial role in the regulation of gene expression during development and differentiation. Characterized by the simultaneous 
+presence of both activating (H3K4me3) and repressive (H3K27me3) histone modifications, bivalent promoters mark genes that 
+are poised for transcription but are not actively transcribed. This dual modification serves as a regulatory mechanism, 
+ensuring that genes essential for differentiation and development are ready to be activated at the appropriate time. 
+Bivalent domains are predominantly found in embryonic stem cells and are crucial for maintaining the cells in a 
+pluripotent state, allowing for the rapid activation or repression of gene expression in response to developmental 
+cues. The significance of bivalent promoters extends to their role in cell fate decisions, where they contribute 
+to the tight control of developmental pathways and the maintenance of stem cell identity\cite{Bernstein2006-wt,Voigt2013-fe}. 
+Our results suggest that the bivalent promoters we identified in blood are inactive (as seen by the fact that the proximal genes are not expressed). 
+However, the fact that DNA methylation at these sites increases with age suggests that they may be losing H3K4me3 with age. H3K4me3 is a critical 
+regulator of DNA methylation as it inhibits the binding of DNMT3 to histones, as the DNMT3 ADD domain preferentially binds to the unmethylated 
+H3K4 residue\cite{Ooi2007-dw}.  This explains why promoters, which are enriched for H3K4me3, are generally hypomethylated.  Our results 
+suggests that there must therefore be an age associated loss of H3K4me3 at these bivalent promoters. That is in fact what we saw when we
+ examined these marks in B cells and Nk cells of both young and old individuals.  These mechanisms further suggest that the age 
+ associated DNA methylation increases may not have a functional consequence in blood and that their proximal genes remain repressed throughout life.
 
-MSEPM model training begins with an  input matrix of $i$ methylation sites for $j$ individuals 
-and $k$ input traits for $j$ individuals. The goal of the MSEPM is to minimize the difference 
-between the observed and predicted methylation values at any of the $i$ input methylation sites as 
-explained by the $k$ input traits. The MSEPM model is fit as follows:
-\begin{itemize}
-
-    \item{Fit $i$ ordinary least squares regression models with the 
-    form $\hat{m_{i,j}} = r^0_i + \sum^n_{k=1} s_{j,k} r_{i,k} + \epsilon_{i,j}$ to estimate site 
-    specific parameters $r_i$ shared amongst all individuals. }
-    \begin{itemize}
-        \item{The $k$ untransformed input traits are used as the initial guess of $S_{j,k}$ 
-        for the first model iteration. }
-    \end{itemize}
-    \item{Update $S_{j,k}$ to minimize the RSS cost 
-    function, $J(k) = \frac{1}{n}\sum_{i=1}^n(\hat{m_{ij}} - (r^0_i + R_i S_j))^2)$, using 
-    gradient descent with fixed $r_{i}$ site parameters.}
-        \begin{itemize}
-            \item{$\nabla J(k)$}
-            \begin{itemize}
-                \item{ $\frac{d}{dk} = \frac{2}{n} \sum_{i=1}^n (\hat{m_{ij}} - (r^0_i + R_i S_j)) r_{i,k}$}
-                \item{$s^1_{j,k} = s^0_{j,k} - \eta * \frac{d}{dk}$, $\eta=$ learning rate}
-            \end{itemize}
-        \end{itemize}
-    \item{Repeat steps 1-2 until the reduction in sum of site RSS between iterations is $\leq$ 
-    specified threshold or a set number of iterations has been reached. }
-\end{itemize}
-
-\subsection{Simulation Framework}
-
-Under the MSEPM formulation a single site can be described linearly where the observed methylation value is 
-dependent on a weighted sum of $n$ epigenetic states, 
-$\hat{m_{i,j}} = r^0_i + \sum^n_{n=1} p_{j,n} r_{i,n} + \epsilon_{i,j}$, where $p_{j,n}$ is an individual 
-trait associated with the methylation status at site $i$. Using this formulation we implemented a simulation 
-framework in python with only numpy \cite{Harris2020-yb} as a dependency.  The details of the approach are as follows:
-
-\begin{itemize}
-    \item{Individuals are simulated by setting individual traits with the following form 
-    $p_{n,j} = Age_j^{\gamma_{n}} q_{n,j}$ Where the simulated individual trait $p_{n,j}$ is 
-    dependent time and exposure, $q_{n,j}$, over time. Under this formulation traits may have a 
-    nonlinear association with the simulated methylation status as mediated by time. The form of  $Q[q]$ varies 
-    between simulated traits as shown above. Simulated MSEPM models are fit using $q_{n,j}$ as input for and the 
-    association between $p_{n,j}$ is learned through model fitting.$p_{n,j} = Age_j^{\gamma_{n}} q_{n,j}$ Where 
-    the simulated individual trait $p_{n,j}$ is dependent time and exposure, $q_{n,j}$, over time. Under this 
-    formulation traits may have a nonlinear association with the simulated methylation status as mediated by time. 
-    The form of  $Q[q]$ varies between simulated traits as shown above. Simulated MSEPM models are fit
-    using $q_{n,j}$ as input for and the association between $p_{n,j}$ is learned through model fitting.}
-    \item{Methylation site profiles are then simulated by setting the dynamic range of possible methylation values 
-    and the error profile of the methylation site. The simulated methylation values are bounded by 0 and 1, and 
-    represent the proportion of methylated bases to the total number of observed cytosines at any site. The 
-    dynamic range is set by first initial methylation value $m_0 ~ \mathcal{beta}(.2,.2)$ and then setting a 
-    target methylation value $m_t$ where the dynamic range is $\delta_i = m_t - m_o$. The value of $m_t$ is 
-    set conditionally to ensure dynamic range is always larger than some specified threshold ,$\theta$, where 
-    $ \theta \leq|\delta| \geq$. $m_0 ~ \mathcal{beta}(.2,.2)$ and then setting a target methylation value 
-    $m_t$ where the dynamic range is $\delta_i = m_t - m_o$. The value of $m_t$ is set conditionally to 
-    ensure dynamic range is always larger than some specified threshold ,
-    $\theta$, where $ \theta \leq|\delta| \geq$. }
-    \item{Methylation sites profiles are then randomly selected and associated with a weighted set of simulated 
-    traits. The weighted traits are then normalized so the input combination of traits describes the dynamic 
-    range of the simulation site, $\delta$. If no weights are provided the individual methylation sites are 
-    simulated by returning $m_0 + N(\mu, \sigma)$. $\delta$. If no weights are provided the individual 
-    methylation sites are simulated by returning $m_0 + N(\mu, \sigma)$. }
-\end{itemize}
-
-\subsection{Blood MSEPM Model Training}
-
-Methylation sites with an absolute pearson correlation coefficient greater than 0.7, 0.995, 0.92 and 0.64 between 
-age $(n=276)$, sex $(n=49)$, cell type PC1 $(n=120)$ and cell type PC2 $(n=238)$ and site methylation values 
-respectively among training samples samples($n=1203$) were included in the model. Additionally, sites with a 
-sum of all absolute pearson coefficients greater than 1.8 were also included $(n=238)$ for a total of 778 
-methylation sites. Min-max, (0-1), scalers were fit using the training input features. Validation and testing 
-sample features were transformed with the trained scalers, note age was min-max scaled on a range from 0-100 
-years. MSEPM models were trained with a learning rate of 0.01 with an iteration limit of 200.
-
-\subsection{Blood MSEPM Model Cluster Transcription Factor Overlap Analysis}
-
-A custom transcription factor reference set was built from all ENCODE V4 transcription factor chromatin 
-immunoprecipitation \cite{ENCODE_Project_Consortium2012-oe,Davis2018-ha} (release 1.4.0 - 2.1.2)  
-irreproducible discovery rate narrow bed peaks, which contains peaks with high rank consistency between 
-replicates, that were not audited for non-compliance or errors. GRCh38 region coordinates were 
-lifted to GRCh37 coordinates using liftOver \cite{Hinrichs2006-oq}. The overlap reference contains 714 
-transcription factor targets 1621 accession IDs (table …).
-
-Blood MSEPM model site hierarchical clustering was performed as follows. Individual methylation site coefficients 
-were first normalized by the standard deviation of methylation values of the site among the training samples, 
-$r_{i,n} / \sigma_{i}$. A distance matrix was then created by taking the Euclidean distance between the 
-normalized site model coefficients. Sites were then clustered using Ward's method which seeks to minimize 
-within cluster variance by minimizing the increase in the error sum of squares (ESS) through successive 
-cluster fusions.  Clusters label by tree cutting at a height of 18. All clustering analysis was carried out 
-using SciPy v1.6.3 \cite{Virtanen2020-mm}.  $r_{i,n} / \sigma_{i}$. A distance matrix was then created by 
-taking the Euclidean distance between the normalized site model coefficients. Sites were then clustered using 
-Ward’s method which seeks to minimize within cluster variance by minimizing the increase in the error sum of 
-squares (ESS) through successive cluster fusions.  Clusters label by tree cutting at a height of 18. All 
-clustering analysis was carried out using SciPy v1.6.3 (Virtanen et al. 2020).
-
-Transcription factor enrichment analysis was performed with  LOLA \cite{Sheffield2016-wg} which assesses the genomic 
-region set overlap between a set of query regions and a set of reference regions, within a specified shared 
-background set, using Fisher's exact test. Overlap analysis was performed for sites within a cluster against 
-the ENCODE V4 reference region (1BP minimum overlap) using all sites assayed with 
-Infinium HumanMethylation450K BeadChip as background.
-
-\subsection{Analysis Environment}
-
-Analysis was carried out in a Jupyter\cite{Basu_undated-vq} analysis environment. 
-Joblib\cite{Varoquaux2009-al}, SciPy\cite{Virtanen2020-wt}, Matplotlib\cite{Hunter2007-nq}, 
-Seaborn\cite{Waskom2021-gj}, Pandas\cite{McKinney2012-ta} and TQDM\cite{Da_Costa-Luis2019-lr} 
-packages were utilized during analysis.
+In conclusion, we introduced a multi-dimensional extension of the Epigenetic Pacemaker, the MSEPM. The MSEPM is capable of 
+accurately modeling multiple methylation associated factors simultaneously. This paradigm can elucidate the site specific 
+regulation underpinning methylome dynamics. It allows us to characterize the mechansims underlying age assocated increases 
+in methylation sites, suggeting that these were caused by the loss of H3K4Me3 at bivalent promoters of genes that are silenced 
+in blood.  The MSEPM is available under the MIT license at https://github.com/NuttyLogic/MultistateEpigeneticPacemaker.
 
@@
 
+\section{Acknowledgments}
+This work has benefited from the equipment and framework of the COMP-HUB and COMP-R Initiatives, funded by 
+the ‘Departments of Excellence’ program of the Italian Ministry for University and Research (MIUR, 2018-2022 and MUR, 2023-2027).
+
+\section{Ethical Statement/Conflict of Interest}
+We have no conflicts of interest to disclose.
+
 \end{linenumbers}
@@
 \printbibliography
+
+\newpage
 
@@
     \footnotesize
-    \caption*{\small \textbf{Figure 1:} Simulated factors and the association with simulated methylation values. (A) 
-    Age with a non-linear association with methylation (B). Sex (C) with a binary association with methylation (D). 
-    Normal factor (E) with a linear relationship with methylation (F). Continuous exponential trait (G) with a linear 
-    relationship with methylation. (I) Simulated methylation sites. Each simulation site has a starting methylation 
-    value $r^0_i$, rate of change associated with each simulated factor $r_{i, factor}$ and range of 
-    variation $\delta_i$. }
+    \caption*{\small \textbf{Figure 1:}Simulated factors and the association with simulated methylation values. (A) Age with a non-linear association with methylation 
+(B). Sex (C) with a binary association with methylation (D). Normal factor (E) with a linear relationship with methylation (F). Continuous exponential
+ trait (G) with a linear relationship with methylation. (I) Simulated methylation sites. Each simulation site has a starting methylation value $r^0_i$ ,
+  rate of change associated with each simulated factor $r_{i, factor}$ and range of variation $\delta_i$.}
     \end{figure}
@@
     \footnotesize
-    \caption*{\small \textbf{Figure 2:} (A) The MAE of the factor predictions on the testing set as multiple factors 
-    are modeled simultaneously. (B) Prediction methylation MAE as factors are included in the MSEPM model. (C) 
-    Model coefficients for Age, Sex, Continuous Normal and Continuous Exponential factors for models trained 
-    $(n=500)$ with all four simulated factors. (D). Simulated and predicted methylation values for all simulated 
-    testing sites across all training folds. }
+    \caption*{\small \textbf{Figure 2:} (A) The MAE of the factor predictions on the testing set as multiple factors are modeled simultaneously. (B) Prediction methylation 
+MAE as factors are included in the MSEPM model. (C) Model coefficients for Age, Sex, Continuous Normal and Continuous Exponential factors for models 
+trained $(n = 500)$ with all four simulated factors. (D). Simulated and predicted methylation values for all simulated testing sites across all training folds. }
     \end{figure}
@@
     \footnotesize
-    \caption*{\small \textbf{Figure 3:} Distribution of age (A) and (B) sex in aggregate blood dataset. (C) 
-    Calculated cell type composition and (D) loading plot of principal components of cell 
-    type composition in aggregate blood data set. .}
+    \caption*{\small \textbf{Figure 3:} Distribution of age (A) and (B) sex in aggregate blood dataset. (C) Calculated cell type composition and (D) loading plot of
+ principal components of cell type composition in aggregate blood data set.}
     \end{figure}
@@
 
+\begin{center}
+    \begin{figure}
+    \includegraphics[scale=.15]{Figures/Figure6.png}
+    \footnotesize
+    \caption*{\small \textbf{Figure6: } (A) Heatmap of H3K4me3 ChIP-seq enrichment for two different blood cell types (CD38\textsuperscript{+} B Cells and CD56\textsuperscript{+} NTK Cells) in 
+two cohorts of individual within 0 to 5 years old and 60 to 65 years old. The average level within 2kb up and downstream for centered genomic regions of cluster 7 is 
+represented above the heatmap. (B) Genome browser view of H3K4me3 levels in each cohort at the promoter regions of \textit{KCTD1} and \textit{IRS2} genes.  
+    }
+    \end{figure}
+\end{center}
+
+\begin{center}
+    \begin{figure}
+    \includegraphics[scale=.15]{Figures/Figure7.png}
+    \footnotesize
+    \caption*{\small \textbf{Figure7: } (A) Heatmap of H3K4me3 ChIP-seq enrichment for two different blood cell types (CD38\textsuperscript{+} B Cells and CD56\textsuperscript{+} NTK Cells) in 
+two cohorts of individual within 0 to 5 years old and 60 to 65 years old. The average level within 2kb up and downstream for centered genomic regions of cluster 7 is 
+represented above the heatmap. (B) Genome browser view of H3K4me3 levels in each cohort at the promoter regions of \textit{KCTD1} and \textit{IRS2} genes. }
+    \end{figure}
+\end{center}
+
 \end{document}

### 2024-05-19 Paper/MSEPM-ManuscriptSupplement.tex
Manuscript revision

@@
 identification of combinations of factors that influence DNA methylation - Supplementary Information}}\\ \\
-\noindent{\small Colin Farrell$^{1,3}$, Kalsuda Lapborisuth$^{1}$, 
-                 Sagi Snir$^{2}$,\\ and Matteo Pellegrini$^{1,3}$\\ \\
+\noindent{\small Colin Farrell$^{1,4}$, Keshiv Tandon ${1}$, Roberto Ferrari$^{2}$, Kalsuda Lapborisuth$^{1}$, 
+                 Sagi Snir$^{3}$,\\ and Matteo Pellegrini$^{1,4}$\\ \\
 \noindent{\footnotesize 

### 2024-05-19 Analysis/Figure6_finalClustergram.m

@@ -1,0 +1,9 @@
+simpMatrix = matfile('simpleMatrix.mat');
+simpleMatrix = simpMatrix.simpleMatrix;
+cgo = clustergram(simpleMatrix, 'Cluster', 'Row');
+set(0, 'ShowHiddenHandles', 'on')
+allhnds = get(0, 'Children');
+h = findall(allhnds, 'Tag', 'HeatMapAxes');
+set(h, 'FontSize', 14)
+
+get(cgo)

### 2024-05-19 Analysis/Figure6_finalMatrix.m

@@ -1,0 +1,25 @@
+simpleMatrix = zeros(95, 93);
+for a = 0:94
+    for y = 1:93
+        file = ['dna-methylation-changes (' num2str(y-1) ')'];
+        data = readtable(file);
+        m = 1;
+        while (round(data{m, 1}) ~= a) & (m < height(data))
+            m = m + 1;
+        end
+        if m ~= height(data)
+            n = m;
+            while round(data{n, 1}) == a
+                n = n + 1;
+            end
+            data00 = nanmean(data(m:n, 2:3), 2);
+            simpleMatrix(a + 1, y) = nanmean(table2array(data00(:, 1)));
+        end
+    end
+end
+save('simpleMatrix.mat', 'simpleMatrix');
+simpMatrix = matfile('simpleMatrix.mat');
+simpleMatrix = simpMatrix.simpleMatrix;
+simpleMatrix(simpleMatrix == 0) = NaN;
+simpleMatrix = knnimpute(simpleMatrix);
+save('simpleMatrix.mat', 'simpleMatrix');

### 2024-06-17 Paper/MSEPM-Manuscript.tex
Update MSEPM-Manuscript.tex
minor edit

@@
 continuous exponentially (CE) distributed trait resembling obesity with a linear association with methylation status 
-$$q \sim \frac{1}{20}e^{-x/20}, s_{CE}=q, \text{Figure 1G-H})$$
+$$q \sim \frac{1}{20}e^{-x/20}, s_{CE}=q, \text{Figure 1G-H}$$
 

### 2024-06-20 Paper/MSEPM-Manuscript.tex
Update MSEPM-Manuscript.tex
changes for everything in email besides figure changes and exponential distribution

--- 2024-05-19 Paper/MSEPM-Manuscript.tex
@@
 sites given the $r_{i,k}$ coefficients set for the respective input factor.  These epigenetic factors can then be used to find the expected methylation value 
-using the trained individual site models where $$E[m_{i,j}] = r_{i,0} + P_{j} \dot R_{i}$$
+using the trained individual site models where $$E[m_{i,j}] = r_{i,0} + P_{j} \dot R_{i}$$ where $P_{j} \dot R_{i}$ is a matrix of point values p and r.
 
@@
 
-Methylation sites were simulated by first randomly setting the dynamic range of the methylation site, 
-$-1< \delta < 1$ a site intercept, $r^0_i$, and the site error, $\sigma_i \sim \mathcal{U}(0.025, 0.05)$. The dynamic 
+Methylation sites were simulated by first randomly setting the possible range of the methylation site, 
+$-1< \delta < 1$ a site intercept, $r^0_i$, and the site error, $\sigma_i \sim \mathcal{U}(0.025, 0.05)$. The 
 range of the methylation site is described by the initial methylation value, $m_0 ~ \beta(.2,.2)$, 
-and the target methylation value, $m_t$, where the dynamic range is $\delta_i = m_t - m_o$.
-The value of $m_t$ is set conditionally to ensure dynamic range is always larger than some specified threshold, $\theta$, where 
+and the target methylation value, $m_t$, where the range is $\delta_i = m_t - m_o$. $\beta(.2,.2)$ is the beta distribution with its parameters for randomly setting the sample.
+The value of $m_t$ is set conditionally to ensure possible range is always larger than some specified threshold, $\theta$, where 
 $ \theta \leq|\delta| \geq. r^0_i ~ \beta(.2,.2)$.
@@
 associated with multiple factors were set by sampling from a uniform distribution. The weighted factor rates are normalized so 
-the input combination of traits describes the dynamic range of the simulated site, $\delta$. If a site is associated 
+the input combination of traits describes the range of the simulated site, $\delta$. If a site is associated 
 with no factors the observed methylation status of a site is described by a random normal with a characteristic offset, $\hat{m_i} = r_{0,i} + N(\mu, \sigma)$.

### 2024-06-20 Paper/MSEPM-Manuscript.tex
Update MSEPM-Manuscript.tex
minor changes for distribution explanation

--- 2024-05-19 Paper/MSEPM-Manuscript.tex
@@
 We simulated individuals whose methylation is determined by four factors and their associated epigenetic factors: 
-a uniform factor approximating age with a non-linear association with methylation status 
-$$q \sim\mathcal{U}(0,100), s_{Age}=q^{0.5}, \text{Figure 1A-B}$$ a binary trait resembling sex, linearly associated 
+a uniformly distributed factor approximating age with a non-linear association with methylation status 
+$$q \sim\mathcal{U}(0,100), s_{Age}=q^{0.5}, \text{Figure 1A-B}$$ a binary distributed trait resembling sex, linearly associated 
 with methylation status $$q \sim B(1,.5), s_{Sex}=q, \text{Figure 1C-D}$$ a continuous normal (CN) phenotype 

### 2024-06-23 Paper/MSEPM-Manuscript.tex
Manuscript Revision

--- 2024-06-17 Paper/MSEPM-Manuscript.tex
@@
 $$q \sim\mathcal{U}(0,100), s_{Age}=q^{0.5}, \text{Figure 1A-B}$$ a binary trait resembling sex, linearly associated 
-with methylation status $$q \sim B(1,.5), s_{Sex}=q, \text{Figure 1C-D}$$ a continuous normal (CN) phenotype 
-resembling a cell type with a linear association with methylation status $$q \sim\mathcal{N}(1, 0.1), s_{CN}=q, 
+with methylation status $$q \sim B(1,.5), s_{Sex}=q, \text{Figure 1C-D}$$ a continuous normal (CN) phenotype a linear 
+association with methylation status $$q \sim\mathcal{N}(1, 0.1), s_{CN}=q, 
 \text{Figure 1E-F}$$ and a 
-continuous exponentially (CE) distributed trait resembling obesity with a linear association with methylation status 
+continuous exponentially (CE) distributed trait with a linear association with methylation status 
 $$q \sim \frac{1}{20}e^{-x/20}, s_{CE}=q, \text{Figure 1G-H}$$
@@
     \footnotesize
-    \caption*{\small \textbf{Figure 2:} (A) The MAE of the factor predictions on the testing set as multiple factors are modeled simultaneously. (B) Prediction methylation 
-MAE as factors are included in the MSEPM model. (C) Model coefficients for Age, Sex, Continuous Normal and Continuous Exponential factors for models 
-trained $(n = 500)$ with all four simulated factors. (D). Simulated and predicted methylation values for all simulated testing sites across all training folds. }
+    \caption*{\small \textbf{Figure 2:} (A) The MAE of the factor predictions on the testing set as multiple factors are modeled simultaneously and 
+    (B) predicted methylation MAE as factors are included in the MSEPM model where the centerline is the 50th quantile and the box 
+    with greatest width contains 50\% of the underlying data with each smaller box containing 50\% of the remaining data 
+    with 6 levels of box width. (C) Model coefficients for 
+Age, Sex, Continuous Normal and Continuous Exponential factors for models trained $(n = 500)$ with all four simulated 
+factors. (D). Simulated and predicted methylation values for all simulated testing sites across all training fold }
     \end{figure}
@@
 \caption*{\small \textbf{Figure 4:} MSEPM model trained with age, sex, CT-PC1 and CT-PC2 predictions within testing 
-set for methylation associated factors (A) age, (B) sex, (C) CT-PC1 and (D) CT-PC2. (E) Observed and predicted
- methylation values for training set has high concordance $(R^2=0.833, MAE=0.035)$}
+set for epigenetic factors (A) age, (B) sex, (C) CT-PC1 and (D) CT-PC2. (E) Observed and predicted
+ methylation values for training set has high concordance $(R^2=0.933, MAE=0.035)$}
 \end{figure}

### 2024-06-23 Paper/MSEPM-Manuscript.tex
Update MSEPM-Manuscript.tex
retry to update file

@@
 
-Methylation sites were simulated by first randomly setting the dynamic range of the methylation site, 
-$-1< \delta < 1$ a site intercept, $r^0_i$, and the site error, $\sigma_i \sim \mathcal{U}(0.025, 0.05)$. The dynamic 
+Methylation sites were simulated by first randomly setting the range of the methylation site, 
+$-1< \delta < 1$ a site intercept, $r^0_i$, and the site error, $\sigma_i \sim \mathcal{U}(0.025, 0.05)$. The possible 
 range of the methylation site is described by the initial methylation value, $m_0 ~ \beta(.2,.2)$, 
-and the target methylation value, $m_t$, where the dynamic range is $\delta_i = m_t - m_o$.
-The value of $m_t$ is set conditionally to ensure dynamic range is always larger than some specified threshold, $\theta$, where 
+and the target methylation value, $m_t$, where the range is $\delta_i = m_t - m_o$. $\beta(.2,.2)$ is the beta distribution with its parameters for randomly setting the sample.
+The value of $m_t$ is set conditionally to ensure site variability is always larger than some specified threshold, $\theta$, where 
 $ \theta \leq|\delta| \geq. r^0_i ~ \beta(.2,.2)$.
@@
 We simulated individuals whose methylation is determined by four factors and their associated epigenetic factors: 
-a uniform factor approximating age with a non-linear association with methylation status 
-$$q \sim\mathcal{U}(0,100), s_{Age}=q^{0.5}, \text{Figure 1A-B}$$ a binary trait resembling sex, linearly associated 
+a uniformly distributed factor approximating age with a non-linear association with methylation status 
+$$q \sim\mathcal{U}(0,100), s_{Age}=q^{0.5}, \text{Figure 1A-B}$$ a binary distributed trait resembling sex, linearly associated 
 with methylation status $$q \sim B(1,.5), s_{Sex}=q, \text{Figure 1C-D}$$ a continuous normal (CN) phenotype a linear 

### 2024-06-23 Paper/MSEPM-Manuscript.tex

--- 2024-06-23 Paper/MSEPM-Manuscript.tex
@@
 \noindent{\small Colin Farrell$^{1,4}$, Keshiv Tandon$^{1}$, Roberto Ferrari$^{2}$, Kalsuda Lapborisuth$^{1}$, 
-                 Sagi Snir$^{3}$,\\ and Matteo Pellegrini$^{1,4}$\\ \\
+                 Sagi Snir$^{3}$, Rahil Modi$^{1}$\\ and Matteo Pellegrini$^{1,4}$\\ \\
 \noindent{\footnotesize 

### 2024-06-23 Paper/MSEPM-Manuscript.tex
Update

--- 2024-06-23 Paper/MSEPM-Manuscript.tex
@@
 \noindent{\small Colin Farrell$^{1,4}$, Keshiv Tandon$^{1}$, Roberto Ferrari$^{2}$, Kalsuda Lapborisuth$^{1}$, 
-                 Sagi Snir$^{3}$, Rahil Modi$^{1}$\\ and Matteo Pellegrini$^{1,4}$\\ \\
+                 Rahil Modi$^{1}$, Sagi Snir$^{3}$, and Matteo Pellegrini$^{1,4}$\\ \\
 \noindent{\footnotesize 

### 2024-09-18 Paper/MSEPM-Manuscript.tex
Update MSEPM-Manuscript.tex
more minor edits

@@
  has been linked to reduced skeletal muscle regeneration\cite{Carlson2009-uz,Paris2016-fo} and SMAD4 polymorphisms 
- are associated with longevity\cite{Carrieri2004-by}). REST is a transcriptional repressor of neuron specific genes in 
+ are associated with longevity\cite{Carrieri2004-by}. REST is a transcriptional repressor of neuron specific genes in 
  non-neuronal cells\cite{Chong1995-dj,Coulson2005-pb}. REST expression is upregulated in aged prefrontal cortex 
  tissue and the absence of REST expression is associated with cognitive impairment\cite{Lu2014-dz} and 
- cellular senescence in neurons\cite{Rocchi2021-od}).
+ cellular senescence in neurons\cite{Rocchi2021-od}.
 
@@
 
-Based on these results we hypothesize that the mechanisms that underlies the gain of methylation with age at these 
+Based on these results we hypothesize that the mechanisms that underlie the gain of methylation with age at these 
 bivalent promoters is the age-associated loss of H3K4me3.  It is well established that the presence of trimethylation
@@
 regulation underpinning methylome dynamics. It allows us to characterize the mechansims underlying age assocated increases 
-in methylation sites, suggeting that these were caused by the loss of H3K4Me3 at bivalent promoters of genes that are silenced 
+in methylation sites, suggesting that these were caused by the loss of H3K4Me3 at bivalent promoters of genes that are silenced 
 in blood.  The MSEPM is available under the MIT license at https://github.com/NuttyLogic/MultistateEpigeneticPacemaker.

### 2024-09-18 Paper/MSEPM-Manuscript.tex

--- 2024-06-23 Paper/MSEPM-Manuscript.tex
@@
   enriched for CTCF transcription factor chip peaks, while those that gain methylation over time are associated with bivalent promoters of 
-  genes that are not expressed in blood. These observation suggest mechanisms that underlie age associated methylation changes and suggest 
+  genes that are not expressed in blood. These observations suggest mechanisms that underlie age associated methylation changes and suggest 
   that age associated increases in methylation may not have strong functional consequences on cell states. In conclusion, the MSEPM is 
@@
 
-We implemented a simulation framework using the MSEPM formulation evaluate the performance of the MSEPM model under various conditions.
+We implemented a simulation framework using the MSEPM formulation to evaluate the performance of the MSEPM model under various conditions.
 To simulate the association between methylation status and an observable factor we modeled the 
@@
 \subsection{H3K4me3 enrichment analysis} 
-Enrichment of analysis for H3K4me3 (figure 7A) was carried out by downloading rpm normalized bigwig files of H3K4me3 ChIP-seq data from 
+Enrichment of analysis for H3K4me3 (Figure 7A) was carried out by downloading rpm normalized bigwig files of H3K4me3 ChIP-seq data from 
 epigenomesportal\cite{Bujold2016-vk} for CD38+ B Cells and CD56+ NTK Cells (for both 0-5 years old and 60-65 years old individuals). 
@@
 bigwig signal over genomic regions of cluster 7 as the BED input. The IGV genome browser\cite{Robinson2011-he} was used to generate an image of the KCTD1 
-and IRS2 promoter regions shown in figure 7B using downloaded bigwig tracks.
+and IRS2 promoter regions shown in Figure 7B using downloaded bigwig tracks.
 
@@
  absolute error (MAE) between the predicted epigenetic states and the simulated epigenetic factors decreases (Figure 2A). Importantly, to accurately
-  assess simulated age it is necessary to account for the influence of the other simulated factors (Sex, CN, CE). The MAE between the predicted 
-  and simulated methylation values decreases as simulated factors are added to the model, and accurately assessing the methylation status of a 
-  simulated site requires that the factor associated with the methylation status at the site is included in the model (Figure 2A).
+  assess simulated age it is necessary to account for the influence of the other simulated factors (Sex, CN, CE).
 

### 2024-09-26 Paper/MSEPM-Manuscript-Updated.tex
Paper update

--- 2024-09-18 Paper/MSEPM-Manuscript.tex
@@
-\documentclass{article}
+\documentclass[sn-nature]{sn-jnl}
 \usepackage{latexsym}
@@
 \usepackage{graphicx}
-\usepackage{hyperref}
 \usepackage{placeins}
@@
 \usepackage{subfiles}
-\usepackage[style=nature]{biblatex}
 \usepackage{float}
@@
 \def\X{{\mathcal X }}
-
-\bibliography{MSEPM-Manuscript}
 
@@
 
-\abstract{Epigenetic clocks, DNA methylation based predictive models of chronological age, are often utilized to study aging associated biology. 
+\section{Abstract}\label{abstract}
+
+Epigenetic clocks, DNA methylation based predictive models of chronological age, are often utilized to study aging associated biology. 
 Despite their widespread use, these methods do not account for other factors that also contribute to the variability of DNA methylation data.
@@
   capable of accurately modeling multiple methylation associated factors and the models produced can illuminate site specific 
-  combinations of factors that affect methylation dynamics.}
+  combinations of factors that affect methylation dynamics.
 
@@
  has been linked to reduced skeletal muscle regeneration\cite{Carlson2009-uz,Paris2016-fo} and SMAD4 polymorphisms 
- are associated with longevity\cite{Carrieri2004-by}. REST is a transcriptional repressor of neuron specific genes in 
+ are associated with longevity\cite{Carrieri2004-by}). REST is a transcriptional repressor of neuron specific genes in 
  non-neuronal cells\cite{Chong1995-dj,Coulson2005-pb}. REST expression is upregulated in aged prefrontal cortex 
  tissue and the absence of REST expression is associated with cognitive impairment\cite{Lu2014-dz} and 
- cellular senescence in neurons\cite{Rocchi2021-od}.
+ cellular senescence in neurons\cite{Rocchi2021-od}).
 
@@
 
-Based on these results we hypothesize that the mechanisms that underlie the gain of methylation with age at these 
+Based on these results we hypothesize that the mechanisms that underlies the gain of methylation with age at these 
 bivalent promoters is the age-associated loss of H3K4me3.  It is well established that the presence of trimethylation
@@
 regulation underpinning methylome dynamics. It allows us to characterize the mechansims underlying age assocated increases 
-in methylation sites, suggesting that these were caused by the loss of H3K4Me3 at bivalent promoters of genes that are silenced 
+in methylation sites, suggeting that these were caused by the loss of H3K4Me3 at bivalent promoters of genes that are silenced 
 in blood.  The MSEPM is available under the MIT license at https://github.com/NuttyLogic/MultistateEpigeneticPacemaker.
@@
 
-\printbibliography
+\bibliography{MSEPM-Manuscript-Updated.bib}
 
@@
     \begin{figure}
-    \includegraphics[scale=.25]{Figures/Figure1.png}    
+    \includegraphics[scale=.25]{Figure1.png}    
     \footnotesize
@@
     \begin{figure}
-    \includegraphics[scale=.2]{Figures/Figure2.png}
+    \includegraphics[scale=.2]{Figure2.png}
     \footnotesize
@@
     \begin{figure}
-    \includegraphics[scale=.2]{Figures/Figure3.png}
+    \includegraphics[scale=.2]{Figure3.png}
     \footnotesize
@@
 \begin{figure}
-\includegraphics[scale=.15]{Figures/Figure4.png}
+\includegraphics[scale=.15]{Figure4.png}
 \footnotesize
@@
     \begin{figure}
-    \includegraphics[scale=.15]{Figures/Figure5.png}
+    \includegraphics[scale=.15]{Figure5.png}
     \footnotesize
@@
     \begin{figure}
-    \includegraphics[scale=.15]{Figures/Figure6.png}
+    \includegraphics[scale=.15]{Figure6.png}
     \footnotesize
@@
     \begin{figure}
-    \includegraphics[scale=.15]{Figures/Figure7.png}
+    \includegraphics[scale=.15]{Figure7.png}
     \footnotesize

### 2024-10-04 Paper/MSEPM-Manuscript-Updated.tex
Update MSEPM-Manuscript-Updated.tex

--- 2024-09-26 Paper/MSEPM-Manuscript.tex
@@
  Within each simulation, epigenetic state predictions and methylation site predictions were made for all testing samples. All models captured 
- the nonlinear association between simulated age and methylation (Supp. Figure 1). As the number of factors in the model is increased the mean 
+ the nonlinear association between simulated age and methylation (Supp. Figure 1). As the number of factors in the model is increased, the mean 
  absolute error (MAE) between the predicted epigenetic states and the simulated epigenetic factors decreases (Figure 2A). Importantly, to accurately
@@
   rate of change associated with each simulated factor $r_{i, factor}$ and range of variation $\delta_i$.}
-    \end{figure}
+    \end{}
 \end{center}

### 2024-10-18 Paper/MSEPM-Manuscript-Updated.tex
Update manuscript text

@@
 peripheral blood mononuclear cells). The aggregate data spanned a wide age range (0.0  - 99.0 years, Figure 3A), contained more predicted females ($n=3392$) than males ($n=2295$, Figure 3B) 
-and reasonable predicted cell type abundance estimates (Figure 3C). The first principal component of a PCA modle trained cell type abundance estimates (CT-PC1) is 
+and reasonable predicted cell type abundance estimates (Figure 3C). The first principal component of a PCA model trained cell type abundance estimates (CT-PC1) is 
 largely driven by the relative abundance of granulocytes (Figure 3D), while the second PC (CT-PC2) captures relative 
@@
   rate of change associated with each simulated factor $r_{i, factor}$ and range of variation $\delta_i$.}
-    \end{}
+    \end{figure}
 \end{center}
